### 2022-06-26 manuscript.Rmd
add extended abstract from SDSS

---
title: |
  Teaching Visual Accessibility in the Introductory Data Science Classes: Why, What, When, and How
shorttitle: Teaching Accessibility in Data Science
author:
- name: JooYoung Seo
  affiliation: "1"  # Must be quoted when including comma.
  email: "jseo1005@illinois.edu"  # Only corresponding authors require email.
- name: Mine Dogucu
  affiliation: 2,3

affiliation:
- id: 1
  institution: "School of Information Sciences, University of Illinois Urbana-Champaign, 614 E. Daniel St. Urbana-Champaign, IL, USA"
- id: 2
  institution: "Department of Statistical Science, University College London"
- id: 3
  institution: "Department of Statistics, University of California Irvine"

abstract: |
  Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency. Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and sometimes people with learning disabilities. We argue that instructors need to teach multiple data representation methods so that all students of all abilities can benefit from multi-sensory data augmentation. In this proposed talk, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility. As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we will share specific examples as well as open-access resources that can be utilized by other data science instructors.
keywords: 
  - Data representations
  - Curriculum
  - R

# the following edits should be done by Journal typesetters
page: 1  # set the first page number
month: July
year: 2020
volume: xx
issue: xx
doi: "xx.xxxx/xxxxxxxxx"
received: "April, 2020"  # optional: comment it out if no received date
accepted: "May, 2020"  # optional: comment it out if no accepted date

bibliography: JDSbib.bib
output:
  rticles::jds_article: default
  bookdown::pdf_book:
    base_format: rticles::jds_article # for using bookdown features like \@ref()
---

```{r setup, include = FALSE, echo = FALSE, purl = FALSE}
knitr::opts_chunk$set(echo = FALSE)

# deactivate addition of preamble when this format is used with bookdown
options(bookdown.theorem.preamble = FALSE)
```

# Introduction {#sec:intro}

According to LinkedIn's 2017 U.S. Emerging Jobs Report, machine learning engineers and data scientists rank among the top emerging jobs.
With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula and a consensus on what should be included in such curricula has not yet been reached. In the latest version of \textit{Computing Competencies for Undergraduate Data Science Curricula} by ACM Data Science Task Force, accessibility is a recommended content for the data science curricula \cite{danyluk2021computing}.

Despite the recent recommendations, accessibility has not yet found its place in data science course materials.
One main reason for this is that accessibility is often covered in the computer sciences or the digital arts if at all covered.
However, data science is an interdisciplinary field that is taught in various disciplines including computer science, statistics, mathematics, business schools, and digital humanities.
Thus, the resources and conversations around accessibility have not yet extended to the broader data science education community.

In this proposed talk, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we will share specific examples as well as open-access resources that can be utilized by other data science instructors.
We teach R as the main language in our courses thus our examples would be shaped around using R, however, instructors using a different set of language(s) or even teaching language-agnostic courses may still find the content applicable for their own courses.

Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency. Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and sometimes people with learning disabilities. We argue that instructors need to teach multiple data representation methods so that all students of all abilities can benefit from multi-sensory data augmentation.

# Learning Goals {#sec:goals}

When shaping our curricular content we have considered multi data representations and their relation to accessibility. We outline the learning goals of this content in this section for different ways of representing data.

## Data Visualization

Data visualization is the most common representation of data. 
Colors are an important part of accessibility of data visualizations. 
In our courses, while teaching colors in data visualizations, we introduce color-blindness through simulators. 
Those without any color vision deficiencies experience what data visualizations would look like to a sighted person that does have color blindness. 
Students get to experience this for red-, green-, and blue-blindness. 
For their data visualizations, instead of relying on the default color selection of R packages, students are introduced to the Okabe-Ito color palette which is known to be accessible to people with color vision deficiencies [@okabe]. 
In addition we rely on `scale_fill_viridis_d()` for generating  `ggplot` visualizations that can be perceived by those with common forms of color blindness.

## Data verbalization (i.e., alt text)

Our definition of "data verbalization" refers to a way of representing data patterns using alternative text description (i.e., alt text). Images without alt text markup are not readily accessible to assistive technologies (i.e., screen readers; refreshable braille displays; text-based voice commands) and web search parsers. In other words, appropriately designed alt text can not only give assistive technology users minimum access to visualized data, but also provide all students with a much richer context.

There are two ways to create alt text: (1) auto-generated alt text; and (2) manual alt text.

Currently, `BrailleR::VI()` function has a capability to auto-generate alt text for basic R graphics and ggplot objects. Supported graphic types include as follows: `graphics::hist()`, `graphics::boxplot()`, `ggplot2::geom_bar()`, `ggplot2::geom_histogram()`, and `ggplot2::geom_boxplot()`.

For other types of graphics that  `BrailleR` doesn’t support, we ask our students to manually create an alt text. We rely on Amy Cesal's “simple formula for writing alt text for data visualization” [@cesal] , which is as follows: “alt= _Chart type_ of _type of data_ where _reason for including chart_.” She also provides a great example based on this formula: “Bar chart of gun murders per 100,000 people where America’s murder rate is 6 times worse than Canada, and 30 times Australia.” We highly recommend other instructors try out this alt text formula with their students for manual text description.

Once alt text is prepared either programmatically or manually, it can be embedded in static and dynamic plot objects in R, such as `ggplot2::labs(..., alt = "[alt text]")`, 
`shiny::renderPlot(..., alt = "[alt tex]")`. We also ask our students to provide alt text within their R Markdown chunks by using `fig.alt = "[alt text]"` chunk option, and within Pandoc-flavor/Common Markdown with `![alt text here.](path/to/file)` syntax.

## Data sonification using sound

Our definition of sonification refers to a data representation method using stereo sound and various audible patterns. For example, `plot(x = 1:5, y = 1:5)` can be sonified using `sonify::sonify(x = 1:5, y = 1:5)`. In this example, a scatter plot is represented by a sound. Position on the X axis is communicated using the left-to-right stereo effect; the pitch of the sound indicates position on the Y axis.

Data sonification is very useful when representing linear patterns; however, there is much room for improvement in terms of complex data sonification as of this writing. Thus, we suggest using sonification for simple linear regression models.

## Data tactualization using Swell Form machine

Data tactulization refers to making data visualization in a tactile form so that it can be touchable by fingers. Although this requires some expensive embossing hardware (e.g., braille embosser or Swell Form Heating Machine), this may be one of the most effective and accessible data representations for those who are blind or visually impaired. One of the authors in this talk, developed a simple R package, called `tactileR` that can generate a ready-to-emboss pdf file from R graphics and ggplot objects. Students and instructors can produce their own tactile graphics by using the following functions: `tactileR::brl_begin()`; [their graphic object]; `tactileR::brl_end()`.

# Delivery of Content

We believe that a natural connection between data science and accessibility can be achieved in the data science classroom while presenting data visualization and then introducing the  other aforementioned forms of data representation.
While delivering the content, instructors should try to provide as many opportunities as possible for students to experience using these tools themselves. 
For instance, each student should use a screen reader at least once regardless of their visual ability status and hear the alternate text that they have written.

Although, we have mainly focused on tools and different representations of data, in order to have students develop accessible content, a more crucial accessibility feature is reproducible workflows. 
Scholars often focus on reproducibility and replication of scientific findings. 
We believe that reproducibility can often serve as an accessibility feature in data science context. 
A reproducible workflow that is open-source provides an opportunity for users who are blind or have visual impairment an opportunity to examine data in their preferred form of data representations.
Thus in addition to teaching different representations of data, instructors may consider teaching reproducibility as well.

# Closing Remarks
We firmly believe that an accessible future in the field of data science requires educators to deliberately teach accessibility as part of the curriculum. We also acknowledge that what we have presented here is only a small portion of the intersection of visual accessibility and data science. One of the co-authors was supported through a Teach Access grant to develop curricular materials on teaching accessibility as part of the data science curriculum. These open-access teaching materials can be accessed at [HIDDEN FOR REVIEW].

# References {-}

### 2022-07-03 manuscript.Rmd
fix yaml affiliation
one sentence per line
use md-style italics

@@
 - name: Mine Dogucu
-  affiliation: 2,3
+  affiliation: "2,3"
 
@@
 With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula and a consensus on what should be included in such curricula has not yet been reached.
-In the latest version of \textit{Computing Competencies for Undergraduate Data Science Curricula} by ACM Data Science Task Force, accessibility is a recommended content for the data science curricula \cite{danyluk2021computing}.
+In the latest version of *Computing Competencies for Undergraduate Data Science Curricula* by ACM Data Science Task Force, accessibility is a recommended content for the data science curricula \cite{danyluk2021computing}.
 
@@
 For other types of graphics that  `BrailleR` doesn’t support, we ask our students to manually create an alt text.
-We rely on Amy Cesal's “simple formula for writing alt text for data visualization” [@cesal] , which is as follows: “alt= _Chart type_ of _type of data_ where _reason for including chart_.” She also provides a great example based on this formula: “Bar chart of gun murders per 100,000 people where America’s murder rate is 6 times worse than Canada, and 30 times Australia.” We highly recommend other instructors try out this alt text formula with their students for manual text description.
+We rely on Amy Cesal's “simple formula for writing alt text for data visualization” [@cesal] , which is as follows: “alt= _Chart type_ of _type of data_ where _reason for including chart_.”
+She also provides a great example based on this formula: “Bar chart of gun murders per 100,000 people where America’s murder rate is 6 times worse than Canada, and 30 times Australia.” We highly recommend other instructors try out this alt text formula with their students for manual text description.
 
@@
 For example, `plot(x = 1:5, y = 1:5)` can be sonified using `sonify::sonify(x = 1:5, y = 1:5)`.
-In this example, a scatter plot is represented by a sound.
-Position on the X axis is communicated using the left-to-right stereo effect; the pitch of the sound indicates position on the Y axis.
+In this example, a scatter plot is represented by a sound. Position on the X axis is communicated using the left-to-right stereo effect; the pitch of the sound indicates position on the Y axis.
 

### 2022-07-04 manuscript.Rmd
tidy up references

@@
 
-bibliography: JDSbib.bib
+bibliography: "bib/references.bib"
 output:
@@
 With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula and a consensus on what should be included in such curricula has not yet been reached.
-In the latest version of *Computing Competencies for Undergraduate Data Science Curricula* by ACM Data Science Task Force, accessibility is a recommended content for the data science curricula \cite{danyluk2021computing}.
+In the latest version of *Computing Competencies for Undergraduate Data Science Curricula* by ACM Data Science Task Force, accessibility is a recommended content for the data science curricula @danyluk2021computing.
 

### 2022-07-23 manuscript.Rmd
some house-cleaning

--- 2022-07-03 manuscript.Rmd
@@
   Teaching Visual Accessibility in the Introductory Data Science Classes: Why, What, When, and How
-shorttitle: Teaching Accessibility in Data Science
+shorttitle: Teaching Accessibility in Data Science # nolint # nolint
 author:
@@
   Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency.
-Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and sometimes people with learning disabilities.
+Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and people with learning disabilities.
 We argue that instructors need to teach multiple data representation methods so that all students of all abilities can benefit from multi-sensory data augmentation.
-In this proposed talk, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
+In this paper, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
 As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we will share specific examples as well as open-access resources that can be utilized by other data science instructors.
@@
 
-bibliography: JDSbib.bib
+bibliography: ["bib/references.bib", "bib/packages.bib"]
 output:
@@
 
-```{r setup, include = FALSE, echo = FALSE, purl = FALSE}
+```{r setup, include=FALSE, echo=FALSE, purl=FALSE}
 knitr::opts_chunk$set(echo = FALSE)
@@
 
-# Introduction {#sec:intro}
+```{r pkg, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, purl=FALSE}
+# Load required packages
+library(ggplot2)
+library(scales)
+library(BrailleR)
+library(sonify)
+library(tactileR)
+library(tidyverse)
+```
+
+# Introduction {#intro}
 
 According to LinkedIn's 2017 U.S. Emerging Jobs Report, machine learning engineers and data scientists rank among the top emerging jobs.
-With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula and a consensus on what should be included in such curricula has not yet been reached.
-In the latest version of *Computing Competencies for Undergraduate Data Science Curricula* by ACM Data Science Task Force, accessibility is a recommended content for the data science curricula \cite{danyluk2021computing}.
+With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula, and a consensus on what should be included in such curricula has not yet been reached.
+In the latest version of *Computing Competencies for Undergraduate Data Science Curricula* by ACM Data Science Task Force, accessibility is a recommended content for the data science curricula [@danyluk2021computing].
 
@@
 
-In this proposed talk, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
+In this paper, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
 As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we will share specific examples as well as open-access resources that can be utilized by other data science instructors.
-We teach R as the main language in our courses thus our examples would be shaped around using R, however, instructors using a different set of language(s) or even teaching language-agnostic courses may still find the content applicable for their own courses.
+We teach R [@R-base] as the main language in our courses thus our examples would be shaped around using R, however, instructors using a different set of language(s) or even teaching language-agnostic courses may still find the content applicable for their own courses.
 
@@
 
-# Learning Goals {#sec:goals}
+# Learning Goals {#goals}
 
@@
 
-Currently, `BrailleR::VI()` function has a capability to auto-generate alt text for basic R graphics and ggplot objects. 
+Currently, `BrailleR::VI()` [@R-BrailleR] function has a capability to auto-generate alt text for basic R graphics and ggplot objects. 
 Supported graphic types include as follows: `graphics::hist()`, `graphics::boxplot()`, `ggplot2::geom_bar()`, `ggplot2::geom_histogram()`, and `ggplot2::geom_boxplot()`.
@@
 Our definition of sonification refers to a data representation method using stereo sound and various audible patterns.
-For example, `plot(x = 1:5, y = 1:5)` can be sonified using `sonify::sonify(x = 1:5, y = 1:5)`.
+For example, `plot(x = 1:5, y = 1:5)` can be sonified using `sonify::sonify(x = 1:5, y = 1:5)` [@R-sonify].
 In this example, a scatter plot is represented by a sound. Position on the X axis is communicated using the left-to-right stereo effect; the pitch of the sound indicates position on the Y axis.
@@
 Although this requires some expensive embossing hardware (e.g., braille embosser or Swell Form Heating Machine), this may be one of the most effective and accessible data representations for those who are blind or visually impaired.
-One of the authors in this talk, developed a simple R package, called `tactileR` that can generate a ready-to-emboss pdf file from R graphics and ggplot objects. 
+One of the authors in this paper, developed a simple R package, called `tactileR` [@R-tactileR] that can generate a ready-to-emboss pdf file from R graphics and ggplot objects. 
 Students and instructors can produce their own tactile graphics by using the following functions: `tactileR::brl_begin()`; [their graphic object]; `tactileR::brl_end()`.
@@
 # References {-}
+
+```{r include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="pkg"}
+knitr::write_bib(file = "bib/packages.bib")
+```

### 2022-07-23 manuscript.Rmd
annotation

@@
 # Load required packages
+## Any packages loaded in this chunk will automatically be added to `bib/packages.bib` for citation use
 library(ggplot2)

### 2022-07-23 manuscript.Rmd
still house-keeping

@@
 
-Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency. 
+Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency (cite). 
 Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and sometimes people with learning disabilities. 
@@
 For their data visualizations, instead of relying on the default color selection of R packages, students are introduced to the Okabe-Ito color palette which is known to be accessible to people with color vision deficiencies [@okabe].
-In addition we rely on `scale_fill_viridis_d()` for generating `ggplot` visualizations that can be perceived by those with common forms of color blindness.
+In addition we rely on `ggplot2::scale_fill_viridis_d()` for generating `ggplot` visualizations [@ggplot22016] that can be perceived by those with common forms of color blindness.
 
@@
 # Closing Remarks
+
 We firmly believe that an accessible future in the field of data science requires educators to deliberately teach accessibility as part of the curriculum. 

### 2022-07-24 manuscript.Rmd
add some citations

@@
 library(tidyverse)
+library(shiny)
+library(knitr)
 ```
@@
 
-Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency (cite). 
-Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and sometimes people with learning disabilities. 
+Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency [@kimAccessibleVisualizationDesign2021]. 
+Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and sometimes people with learning disabilities [@kimAccessibleVisualizationDesign2021; @marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
 We argue that instructors need to teach multiple data representation methods so that all students of all abilities can benefit from multi-sensory data augmentation.
@@
 Once alt text is prepared either programmatically or manually, it can be embedded in static and dynamic plot objects in R, such as `ggplot2::labs(..., alt = "[alt text]")`, 
-`shiny::renderPlot(..., alt = "[alt tex]")`. 
-We also ask our students to provide alt text within their R Markdown chunks by using `fig.alt = "[alt text]"` chunk option, and within Pandoc-flavor/Common Markdown with `![alt text here.](path/to/file)` syntax.
+`shiny::renderPlot(..., alt = "[alt tex]")` [@R-shiny].
+We also ask our students to provide alt text within their R Markdown chunks by using `fig.alt = "[alt text]"` `knitr` chunk option [@knitr2014], and within Pandoc-flavor/Common Markdown with `![alt text here.](path/to/file)` syntax [@macfarlanePandoc2022].
 

### 2022-07-24 manuscript.Rmd
house-cleaning

@@
 library(knitr)
+library(gt)
 ```
 
-# Introduction {#intro}
+# Introduction
 
@@
 
-# Learning Goals {#goals}
+# Learning Goals
 

### 2022-07-24 manuscript.Rmd
add chunk name
expand data viz section

@@
 output:
-  rticles::jds_article: default
+  rticles::jds_article: 
+    includes:
+      in_header: header.tex #header-hidden.tex can be used when creating an anonymized version for review
   bookdown::pdf_book:
@@
 library(gt)
+library(colorblindr)
+library(palmerpenguins)
 ```
@@
 
-Data visualization is the most common representation of data. 
-Colors are an important part of accessibility of data visualizations. 
-In our courses, while teaching colors in data visualizations, we introduce color-blindness through simulators. 
-Those without any color vision deficiencies experience what data visualizations would look like to a sighted person that does have color blindness. 
-Students get to experience this for red-, green-, and blue-blindness. 
+Data visualization is the most common representation of data and it is covered in many introductory data science curricula and textbooks <!--Mine will add citations-->.
+Despite their use by many, data visualizations are not accessible to all. 
+Colors are an important part of data visualizations and an aspect that can limit accessibility for color blind people. 
+When teaching data visualizations we include three basic rules for incorporating colors into data visualization. Students should be able to
+
+1. simulate color blindness (if they are sighted and not color blind);
+2. pick color blind friendly colors.
+3. not only rely on color for differentiating data;
+
+We show an example of incorporating these three basic rules by using the Palmer Penguins data <!--Mine will cite this-->. 
+We visualize the relationship between flipper length and bill length of penguins and color the data points differently for each species in Figure \ref{fig:penguins-basic}.
+
+```{r penguins-basic, echo=FALSE, fig.cap="Visualization showing the relationship between flipper length (mm) and bill length (mm) of penguins for three different species of penguins", fig.alt="", warning=FALSE, fig.height=3}
+library(ggplot2) #It seems like the pkg chunk is not loading packages. I did not want to make changes there since I do not know what your intention was.
+library(palmerpenguins)
+fig <- ggplot(penguins, aes(flipper_length_mm, bill_length_mm, color = species)) +
+         geom_point() +
+  labs(x = "Flipper Length(mm)",
+       y = "Bill Length (mm)",
+       color = "Species") +
+  theme_bw()
+
+fig
+
+```
+
+The `colorblindr` package provides a set of functions that can be useful in teaching and practising color blindness accessibility. 
+For instance, `cvd_grid()` function provides color-deficiency simulations of a given plot. 
+In the code chunk below the plot is named as `fig`. 
+The code creates a grid of four plots where each plot shows a different color vision deficiency.
+This grid allows those without any color vision deficiencies experience what data visualizations would look like to a sighted person that does have color blindness.
+
+```{r penguins-cvd, echo=FALSE, fig.cap="", fig.alt="", warning=FALSE}
+colorblindr::cvd_grid(fig)
+```
+
 For their data visualizations, instead of relying on the default color selection of R packages, students are introduced to the Okabe-Ito color palette which is known to be accessible to people with color vision deficiencies [@okabe].
-In addition we rely on `ggplot2::scale_fill_viridis_d()` for generating `ggplot` visualizations [@ggplot22016] that can be perceived by those with common forms of color blindness.
+One can find specific colors in the pallette by the R code `palette.colors(palette = "Okabe-Ito")`.
+In addition, students can be introduced to `colorblindr::scale_color_OkabeIto()` and `colorblindr::scale_fill_OkabeIto()` to automatically switch to using this color pallette by adding a layer to a ggplot presentation.
+For instance, the plot in Figure \ref{fig:penguins-basic} can be switched to Okabe-Ito color pallette as shown in Figure \ref{fig:penguins-okabe-ito} by using the code below.
+
+```{r penguins-okabe-ito, fig.cap="", fig.alt="", warning=FALSE, fig.height=3}
+fig + colorblindr::scale_color_OkabeIto()
+```
+
+Last but not least, students should learn not to rely only on color to differentiate data points, in this case Species. 
+For instance, they can be introduced to shapes and have data points that are circle, triangle, square for each species. 
+This can also be extended to line graphs as having continuous or dashed lines.
+Facetting can also be utilized. 
+For instance, each of the species could have its own scatterplot side-by-side.
+
+In this manuscript and in our introductory courses, we focus on static data visualizations. 
+Those who teach interactive data visualizations may also consider covering cognitive and motor accessibility. 
+Our focus here is on visual accessibility but it is worth noting that all subdomains of accessibility are relevant to data science learners and can be incorporated into the curriculum as seen fit.
 
@@
 
-```{r include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="pkg"}
+```{r pkg_bib, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="pkg"}
 knitr::write_bib(file = "bib/packages.bib")

### 2022-07-24 manuscript.Rmd
whitespace

--- 2022-07-24 manuscript.Rmd
@@
 
-According to LinkedIn's 2017 U.S. Emerging Jobs Report, machine learning engineers and data scientists rank among the top emerging jobs.
-With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula, and a consensus on what should be included in such curricula has not yet been reached.
+According to LinkedIn's 2017 U.S. Emerging Jobs Report, machine learning engineers and data scientists rank among the top emerging jobs (cite).
+With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula.
+However, a consensus on what should be included in such curricula has not yet been reached.
 In the latest version of *Computing Competencies for Undergraduate Data Science Curricula* by ACM Data Science Task Force, accessibility is a recommended content for the data science curricula [@danyluk2021computing].

### 2022-07-24 manuscript.Rmd
linting + citation

@@
 ```{r penguins-basic, echo=FALSE, fig.cap="Visualization showing the relationship between flipper length (mm) and bill length (mm) of penguins for three different species of penguins", fig.alt="", warning=FALSE, fig.height=3}
-library(ggplot2) #It seems like the pkg chunk is not loading packages. I did not want to make changes there since I do not know what your intention was.
+library(ggplot2) # It seems like the pkg chunk is not loading packages. I did not want to make changes there since I do not know what your intention was.
 library(palmerpenguins)
@@
   geom_point() +
-  labs(x = "Flipper Length(mm)",
+  labs(
+    x = "Flipper Length(mm)",
     y = "Bill Length (mm)",
-       color = "Species") +
+    color = "Species"
+  ) +
   theme_bw()
@@
 fig
-
 ```
 
-The `colorblindr` package provides a set of functions that can be useful in teaching and practising color blindness accessibility. 
+The `colorblindr` package [@R-colorblindr] provides a set of functions that can be useful in teaching and practising color blindness accessibility. 
 For instance, `cvd_grid()` function provides color-deficiency simulations of a given plot. 

### 2022-07-24 manuscript.Rmd
expand content delivery and closing remarks

--- 2022-07-24 manuscript.Rmd
@@
 
-According to LinkedIn's 2017 U.S. Emerging Jobs Report, machine learning engineers and data scientists rank among the top emerging jobs (cite).
-With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula.
-However, a consensus on what should be included in such curricula has not yet been reached.
+According to LinkedIn's 2017 U.S. Emerging Jobs Report, machine learning engineers and data scientists rank among the top emerging jobs.
+With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula, and a consensus on what should be included in such curricula has not yet been reached.
 In the latest version of *Computing Competencies for Undergraduate Data Science Curricula* by ACM Data Science Task Force, accessibility is a recommended content for the data science curricula [@danyluk2021computing].
@@
 
+Ample opportunities in learning accessibility should also include assessments. 
+For instance, instructors incorporating accessibility should modify assessment instructions and rubrics to include accessibility. 
+This will prevent students just hearing about accessibility in the lecture then forgetting about it. 
+Ideally, the assessments should not treat accessibility as a learning objective at a single point in the academic term. 
+For instance, if accessibility is covered in Week 3 of the term, students can (and should) still be expected to write alternate texts for visualizations in Week 10.
+
+In this manuscript we have focused on domain-specific (i.e. data science) knowledge of accessibility using a particular language (R). 
+Depending on the courses taught, instructors may want to extend the accessibility content to non-data contexts. 
+In addition they may adopt the content in their courses with other languages such as Python and SAS or even in their language-agnostic courses. 
+Overall, if taught well, a student learning accessibility content, for instance, should be able to include alternate text in their PowerPoint presentation in their Anthropology class. 
+In other words, accessibility knowledge should extend beyond data science and beyond any tool.
+
 Although, we have mainly focused on tools and different representations of data, in order to have students develop accessible content, a more crucial accessibility feature is reproducible workflows. 
@@
 We also acknowledge that what we have presented here is only a small portion of the intersection of visual accessibility and data science. 
-One of the co-authors was supported through a Teach Access grant to develop curricular materials on teaching accessibility as part of the data science curriculum.
-These open-access teaching materials can be accessed at [HIDDEN FOR REVIEW].
+The second co-author was supported through a Teach Access grant to develop curricular materials on teaching accessibility as part of the data science curriculum. 
+Educators interested in teaching accessibility can follow the organization's work at https://teachaccess.org/.
 

### 2022-07-24 manuscript.Rmd
comment out in_header: cannot render

--- 2022-07-24 manuscript.Rmd
@@
   rticles::jds_article: 
-    includes:
-      in_header: header.tex #header-hidden.tex can be used when creating an anonymized version for review
+    # includes:
+      # in_header: header.tex #header-hidden.tex can be used when creating an anonymized version for review
   bookdown::pdf_book:

### 2022-07-24 header.tex
add GitHub repo link

% For anonymity
\newcommand{\repo}[1]{https://github.com/mdogucu/teaching-accessibility-manuscript}

% Remove green borders around links
\usepackage{xcolor}
\xdefinecolor{blue}{rgb}{0.28,0.65,0.9}
\hypersetup{
    colorlinks,
    linkcolor={blue},
    citecolor={blue},
    urlcolor={blue}
}

### 2022-07-25 manuscript.Rmd
add GitHub repo link
missing package info

@@
 
-We show an example of incorporating these three basic rules by using the Palmer Penguins data <!--Mine will cite this-->. 
+We show an example of incorporating these three basic rules by using the Palmer Penguins data [@R-palmerpenguins]. 
 We visualize the relationship between flipper length and bill length of penguins and color the data points differently for each species in Figure \ref{fig:penguins-basic}.
@@
 The `colorblindr` package [@R-colorblindr] provides a set of functions that can be useful in teaching and practising color blindness accessibility. 
-For instance, `cvd_grid()` function provides color-deficiency simulations of a given plot. 
+For instance, `colorblindr::cvd_grid()` function provides color-deficiency simulations of a given plot. 
 In the code chunk below the plot is named as `fig`. 

### 2022-07-25 manuscript.Rmd

@@
 We also acknowledge that what we have presented here is only a small portion of the intersection of visual accessibility and data science. 
-One of the co-authors was supported through a Teach Access grant to develop curricular materials on teaching accessibility as part of the data science curriculum.
-These open-access teaching materials can be accessed at [HIDDEN FOR REVIEW].
+The second co-author was supported through a Teach Access grant to develop curricular materials on teaching accessibility as part of the data science curriculum. 
+Educators interested in teaching accessibility can follow the organization's work at https://teachaccess.org/. 
+The source code of this manuscript can be found on a \href{\repo{}}{GitHub repository}^[The GitHub repository can be found at \url{\repo{}}]
 

### 2022-07-25 manuscript.Rmd
merge

@@
   rticles::jds_article: 
-    # includes:
-      # in_header: header.tex #header-hidden.tex can be used when creating an anonymized version for review
+    includes:
+      in_header: header.tex #header-hidden.tex can be used when creating an anonymized version for review
   bookdown::pdf_book:

### 2022-07-25 manuscript.Rmd
add BrailleR::VI() example

@@
 
-Our definition of "data verbalization" refers to a way of representing data patterns using alternative text description (i.e., alt text). 
+Our definition of "data verbalization" refers to a way of representing data patterns using verbal description with alternative text (i.e., alt text). 
 Images without alt text markup are not readily accessible to assistive technologies (i.e., screen readers; refreshable braille displays; text-based voice commands) and web search parsers. 
-In other words, appropriately designed alt text can not only give assistive technology users minimum access to visualized data, but also provide all students with a much richer context.
+However, appropriately designed alt text can not only give assistive technology users minimum access to visualized data, but also provide all students with a much richer context<!--jy will cite-->.
 
@@
 
-Currently, `BrailleR::VI()` [@R-BrailleR] function has a capability to auto-generate alt text for basic R graphics and ggplot objects. 
+Currently, `BrailleR::VI()` [@R-BrailleR] function has a capability to auto-generate alt text for basic R graphics and ggplot objects (see Figure \@ref(fig:VI-example)). 
 Supported graphic types include as follows: `graphics::hist()`, `graphics::boxplot()`, `ggplot2::geom_bar()`, `ggplot2::geom_histogram()`, and `ggplot2::geom_boxplot()`.
+
+```{r VI-example, fig.cap = "Automatic data verbalization.", fig.alt = "Histogram of Ozone with its verbal description automatically generated by `BrailleR::VI()` function.", echo=TRUE, include=TRUE, warning=FALSE, cache=TRUE}
+library(BrailleR)
+BrailleR::VI(hist(airquality$Ozone))
+```
 

### 2022-07-25 manuscript.Rmd
better manage figures in figure directory

@@
 ```{r setup, include=FALSE, echo=FALSE, purl=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, fig.path = "figure/")
 
@@
 
+<!-- ```{r manual-alt, fig.cap = "Figure caption.", fig.alt = "Alt text."} -->
+# knitr::include_graphics("path/to/file.png")
+<!-- ``` -->
+
 Once alt text is prepared either programmatically or manually, it can be embedded in static and dynamic plot objects in R, such as `ggplot2::labs(..., alt = "[alt text]")`, 

### 2022-07-25 manuscript.Rmd
cite linkedin
Clean up introduction

@@
 
-According to LinkedIn's 2017 U.S. Emerging Jobs Report, machine learning engineers and data scientists rank among the top emerging jobs (cite).
+According to LinkedIn's U.S. Emerging Jobs Report, data scientists rank among the top emerging jobs [-@linkedin].
 With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula. 
@@
 Despite the recent recommendations, accessibility has not yet found its place in data science course materials.
-One main reason for this is that accessibility is often covered in the computer sciences or the digital arts if at all covered.
-However, data science is an interdisciplinary field that is taught in various disciplines including computer science, statistics, mathematics, business schools, and digital humanities.
-Thus, the resources and conversations around accessibility have not yet extended to the broader data science education community.
+One main reason for this could be that accessibility is often covered in the computer sciences [@kawas2019teaching] or the digital arts [@barata2019inclusion] if at all covered.
+However, data science is an interdisciplinary field that is taught in various disciplines including computer science, statistics, business schools, political science, and biological sciences [@schwab2021data].
+Thus, we believe that the resources and conversations around accessibility have not yet extended to the broader data science education community at a scale that it should.
+For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on accessibility of data.
 
-In this paper, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
-As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we will share specific examples as well as open-access resources that can be utilized by other data science instructors.
-We teach R [@R-base] as the main language in our courses thus our examples would be shaped around using R, however, instructors using a different set of language(s) or even teaching language-agnostic courses may still find the content applicable for their own courses.
+In this manuscript, we argue that accessibility should be taught early in the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
+As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we share specific examples that can be utilized by other data science instructors.
+We teach R [@R-base] as the main language in our courses thus our examples are shaped around using R, however, instructors using a different set of language(s) or even teaching language-agnostic courses may still find the content applicable for their own courses.
+
+# Learning Goals
 
@@
 We argue that instructors need to teach multiple data representation methods so that all students of all abilities can benefit from multi-sensory data augmentation.
-
-# Learning Goals
 

### 2022-07-25 manuscript.Rmd
extend learning goals

@@
 
-# Learning Goals
+# Learning Goals and Objectives
+
+Both the authors teach data science in the United States thus it was a natural choice for us to make students aware of the Americans with Disabilities Act (ADA). 
+The ADA is a civil rights law that prohibits discrimination of individuals with disabilities. 
+Regardless of their own disability status, as future data scientists, we expect our students to produce data science products that are accessible to all, not only because it is the right thing to do but also because they may have future responsibilities as data scientists to comply to ADA.
+
+Disabled people may rely on different types of assistive technology in their day to day life. 
+Assistive Technology is any form of technology (e.g. software, hardware) that helps people with disabilities perform certain activities.
+Examples of assistive technology include but not limited to walking sticks, wheel chairs and screen readers.
+In incorporating accessibility to our course contents, we wanted students to be familiar with assistive technologies, and since our focus has been on visual accessibility, it was a natural fit to include screen readers.
+A screen reader is an assistive technology that supports blind or visually impaired people in using their computer by reading out the contents. 
+Screen readers are built-in Windows (Settings > Ease of Access > Narrator, and then turn on the toggle under Use Narrator) and Mac OS (System Preferences > Accessibility > Voice Over > Enable Voice Over) and can be used by students.
 
@@
 We argue that instructors need to teach multiple data representation methods so that all students of all abilities can benefit from multi-sensory data augmentation.
+When shaping our curricular content we have considered multi data representations and their relation to accessibility.
 
-When shaping our curricular content we have considered multi data representations and their relation to accessibility.
-We outline the learning goals of this content in this section for different ways of representing data.
+In summary, our curricular goal has been to ensure that the current and the next generation of data scientists, provide public-facing outputs (websites, analysis reports, etc.) that are visually accessible.
+For this goal, we set three specific learning objectives as follows. Students should
+
+1. get familiar with Americans with Disabilities Act;
+2. use at least one assistive technology (i.e. screen reader);
+3. consider different representations of data.
+
+It is important to note that the first two objectives are not data science specific but are related to accessibility in the broader scale. 
+The third objective, however, is data science specific. 
+In the rest of the section we focus on teaching different representations of data further.
 
@@
 
-Data visualization is the most common representation of data and it is covered in many introductory data science curricula and textbooks <!--Mine will add citations-->.
+Data visualization is the most common representation of data.
 Despite their use by many, data visualizations are not accessible to all. 
@@
 1. simulate color blindness (if they are sighted and not color blind);
-2. pick color blind friendly colors.
-3. not only rely on color for differentiating data;
+2. pick color blind friendly colors;
+3. not only rely on color for differentiating data.
 

### 2022-07-25 manuscript.Rmd
clean up color sections

@@
 
-```{r penguins-basic, echo=FALSE, fig.cap="Visualization showing the relationship between flipper length (mm) and bill length (mm) of penguins for three different species of penguins", fig.alt="", warning=FALSE, fig.height=3}
+```{r penguins-basic, echo=FALSE, fig.cap="Visualization showing the relationship between flipper length (mm) and bill length (mm) of penguins for three different species of penguins", fig.alt="A scatterplot showing the relationship between flipper length in mm on the x axis and bill length in mm on the y axis.
+Flipper lengths vary from about 170 to 230 and bill lengths vary from about 35 to 60.
+Overall there is a moderate positive relationship.
+Each data point is colored differently for three species as Adelie, Chindtrap and Gentoo.
+Adelie have short flipper length and short bill length in comparison.
+Chinstrap have low flipper length but high bill length.
+Gentoo have high flipper length and high bill length.
+Data points are in red, green, and blue for Adelie, Chinstrap and Gentoo penguins respectively.", warning=FALSE, fig.height=3}
 library(ggplot2) # It seems like the pkg chunk is not loading packages. I did not want to make changes there since I do not know what your intention was.
@@
 In the code chunk below the plot is named as `fig`. 
-The code creates a grid of four plots where each plot shows a different color vision deficiency.
-This grid allows those without any color vision deficiencies experience what data visualizations would look like to a sighted person that does have color blindness.
+The code creates a grid of four plots where each plot shows a different color vision deficiency including deutanomly, protanomly, tritanomly, and desaturated.
+This grid allows those without any color vision deficiencies experience what data visualizations would look like to a sighted person that has color blindness.
 
-```{r penguins-cvd, echo=FALSE, fig.cap="", fig.alt="", warning=FALSE}
+```{r penguins-cvd, echo=FALSE, fig.cap="Color Blindness Simulation with colorblindr", fig.alt="A 2 by 2 grid of four scatterplots.
+The scatterplots are the same as the scatterplot in the previous figure except for colors.
+From left to right and top to bottom the plots read deutanomly, protanomly, tritanomly, and desaturated", warning=FALSE, echo=TRUE}
 colorblindr::cvd_grid(fig)
@@
 In addition, students can be introduced to `colorblindr::scale_color_OkabeIto()` and `colorblindr::scale_fill_OkabeIto()` to automatically switch to using this color pallette by adding a layer to a ggplot presentation.
-For instance, the plot in Figure \ref{fig:penguins-basic} can be switched to Okabe-Ito color pallette as shown in Figure \ref{fig:penguins-okabe-ito} by using the code below.
+For instance, the plot in Figure \ref{fig:penguins-basic} can be switched to Okabe-Ito color pallette as shown in Figure \ref{fig:penguins-okabe} by using the code below.
 
-```{r penguins-okabe-ito, fig.cap="", fig.alt="", warning=FALSE, fig.height=3}
+```{r penguins-okabe, fig.cap="Scatterplot Using Okabe-Ito Color Pallette", fig.alt="The scatterplots are the same as the scatterplot in the original penguins plot except for colors.
+Data points are in orange, sky blue, and bluish green for Adelie, Chinstrap and Gentoo penguins respectively.", warning=FALSE, fig.height=3, echo=TRUE}
 fig + colorblindr::scale_color_OkabeIto()

### 2022-07-25 manuscript.Rmd
update Rmd and .gitignore

@@
 library(palmerpenguins)
+
 fig <- ggplot(penguins, aes(flipper_length_mm, bill_length_mm, color = species)) +
@@
 
-Currently, `BrailleR::VI()` [@R-BrailleR] function has a capability to auto-generate alt text for basic R graphics and ggplot objects (see Figure \@ref(fig:VI-example)). 
+Currently, `BrailleR::VI()` [@R-BrailleR] function has a capability to auto-generate alt text for basic R graphics and ggplot objects (see Figure \ref{fig:VI-example}). 
 Supported graphic types include as follows: `graphics::hist()`, `graphics::boxplot()`, `ggplot2::geom_bar()`, `ggplot2::geom_histogram()`, and `ggplot2::geom_boxplot()`.
 
-```{r VI-example, fig.cap = "Automatic data verbalization.", fig.alt = "Histogram of Ozone with its verbal description automatically generated by `BrailleR::VI()` function.", echo=TRUE, include=TRUE, warning=FALSE, cache=TRUE}
+```{r VI-example, fig.cap = "Automatic data verbalization.", fig.alt = "Bar chart of penguins species with its verbal description automatically generated by `BrailleR::VI()` function.", echo=TRUE, include=TRUE, warning=FALSE, cache=TRUE}
+library(ggplot2)
 library(BrailleR)
-BrailleR::VI(hist(airquality$Ozone))
+library(palmerpenguins)
+
+# Create graph
+g_species <- ggplot(data = penguins, mapping = aes(x = species)) +
+  geom_bar()
+
+g_species
+
+# Obtain auto verbalization
+# (Added feature in BrailleR >= 0.32.1): `BrailleR::VI()` function is automatically called against supported graphics if BrailleR package is currently loaded in R session
+# BrailleR::VI(g_species)
 ```

### 2022-07-25 manuscript.Rmd
add sonify example

@@
 
-<!-- ```{r manual-alt, fig.cap = "Figure caption.", fig.alt = "Alt text."} -->
-# knitr::include_graphics("path/to/file.png")
-<!-- ``` -->
+<!-- @mine -- Please add example figure with with manual alt text. -->
 
@@
 Our definition of sonification refers to a data representation method using stereo sound and various audible patterns.
-For example, `plot(x = 1:5, y = 1:5)` can be sonified using `sonify::sonify(x = 1:5, y = 1:5)` [@R-sonify].
+For example, Figure \ref{fig:plot-xy} can be made audible by using `sonify` package [@R-sonify] via `r sonify::sonify(x = 1:5, y = 1:5)`.
 In this example, a scatter plot is represented by a sound. Position on the X axis is communicated using the left-to-right stereo effect; the pitch of the sound indicates position on the Y axis.
@@
 Thus, we suggest using sonification for simple linear regression models.
+
+```{r plot-xy, fig.cap="Visualizing the relationship of two numerical variables with scatter plot.", fig.alt="Scatter plot representing the positive relationship between x and y variables.", echo=FALSE}
+# Prepare variables
+x <- 1:5
+y <- 1:5
+
+# Representing the relationship between x and y variables via plot visualization
+plot(x, y)
+```
+
+<!-- ```{r sonify-x-y, echo=TRUE, include=TRUE}
+# Representing the relationship between the two numerical variables via stereo sound
+library(sonify())
+sonify(x, y)
+``` -->
 

### 2022-07-27 manuscript.Rmd
add tactile graph example and code

@@
 One of the authors in this paper, developed a simple R package, called `tactileR` [@R-tactileR] that can generate a ready-to-emboss pdf file from R graphics and ggplot objects. 
-Students and instructors can produce their own tactile graphics by using the following functions: `tactileR::brl_begin()`; [their graphic object]; `tactileR::brl_end()`.
+Students and instructors can produce their own tactile graphics by using the following functions: `tactileR::brl_begin()`; [their graphic object]; `tactileR::brl_end()` (see Figure \ref{fig:tactile-example}).
+
+```{r tactile-producing-code, echo=TRUE, include=TRUE}
+library(palmerpenguins)
+library(tactileR)
+
+brl_begin("figure/tactile-boxplot.pdf")
+with(
+  penguins,
+  boxplot(body_mass_g ~ species)
+)
+brl_end()
+```
+
+```{r tactile-example, fig.cap="Ready-to-print tactile graph in braille.", fig.alt="Boxplot showing the different body mass between penguins species"}
+knitr::include_graphics("figure/tactile-boxplot.pdf")
+```
 

### 2022-07-27 manuscript.Rmd
add sonify code and example

@@
 Our definition of sonification refers to a data representation method using stereo sound and various audible patterns.
-For example, Figure \ref{fig:plot-xy} can be made audible by using `sonify` package [@R-sonify] via `r sonify::sonify(x = 1:5, y = 1:5)`.
+For example, Figure \ref{fig:plot-xy} can be made audible by using `sonify` package [@R-sonify] like below.
+
+```{r sonify-x-y, echo=TRUE, include=TRUE, cache=TRUE}
+# Representing the relationship between the two numerical variables via stereo sound
+library(sonify())
+
+# Prepare variables
+x <- 1:5
+y <- 1:5
+
+sonify(x, y)
+```
+
 In this example, a scatter plot is represented by a sound. Position on the X axis is communicated using the left-to-right stereo effect; the pitch of the sound indicates position on the Y axis.
@@
 
-<!-- ```{r sonify-x-y, echo=TRUE, include=TRUE}
-# Representing the relationship between the two numerical variables via stereo sound
-library(sonify())
-sonify(x, y)
-``` -->
-
 ## Data tactualization using Swell Form machine
@@
 
-```{r tactile-producing-code, echo=TRUE, include=TRUE}
+```{r tactile-producing-code, echo=TRUE, include=TRUE, cache=TRUE}
 library(palmerpenguins)
@@
 
-```{r tactile-example, fig.cap="Ready-to-print tactile graph in braille.", fig.alt="Boxplot showing the different body mass between penguins species"}
+```{r tactile-example, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="tactile-producing-code", fig.cap="Ready-to-print tactile graph in braille.", fig.alt="Boxplot showing the different body mass between penguins species"}
 knitr::include_graphics("figure/tactile-boxplot.pdf")

### 2022-07-27 manuscript.Rmd
add pseudo audio file link for sonify-example.wav
Closes #8

@@
 Our definition of sonification refers to a data representation method using stereo sound and various audible patterns.
-For example, Figure \ref{fig:plot-xy} can be made audible by using `sonify` package [@R-sonify] like below.
+For example, Figure \ref{fig:plot-xy} can be made audible by using `sonify` package [@R-sonify] like below (audio file is available at \url{\audio{}}).
 
@@
 Educators interested in teaching accessibility can follow the organization's work at https://teachaccess.org/. 
-The source code of this manuscript can be found on a \href{\repo{}}{GitHub repository}^[The GitHub repository can be found at \url{\repo{}}]
+The source code of this manuscript can be found on a GitHub repository at \url{\repo{}}.
 

### 2022-07-28 manuscript.Rmd
update learning goals

@@
 
-Both the authors teach data science in the United States thus it was a natural choice for us to make students aware of the Americans with Disabilities Act (ADA). 
+Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency [@kimAccessibleVisualizationDesign2021]. 
+Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and sometimes people with learning disabilities [@kimAccessibleVisualizationDesign2021; @marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
+We argue that instructors need to teach multiple data representation methods so that all students of all abilities can benefit from multi-sensory data augmentation.
+To this end, we have considered multi data representations and their relation to accessibility in our courses, and that's what we will demonstrate throughout this paper.
+
+In incorporating accessibility to our course contents, we also wanted our students to be familiar with assistive technologies, and since our focus has been on visual accessibility, it was a natural fit to include screen readers.
+Screen reader is an assistive technology that supports blind or visually impaired people in using their computer by reading aloud the contents.
+Recently, screen reader is provided as one of the built-in accessibility features on operating systems, such as Windows Narrator and Apple's VoiceOver.
+
+However, we also want to make our teaching as inclusive as possible to other types of dis/abilities so that we are not just confined with issues of visual disabilities.
+
+Therefore, our curricular goal is to broadly ensure that the current and the next generation of data scientists, provide public-facing outputs, including websites and analysis reports, in accessible forms and representations for people with diverse abilities.
+To achieve this goal, we set three specific and minimal learning objectives like below.
+
+Students should:
+
+1. get familiar with Americans with Disabilities Act;
+2. use at least one assistive technology;
+3. consider different representations of data.
+
+The first objective provides our students with a chance to not just learn legislative concepts, but empathy building concerning diversity issue.
+As both the authors teach data science in the United States, it was a natural choice for us to make students aware of the Americans with Disabilities Act (ADA).
 The ADA is a civil rights law that prohibits discrimination of individuals with disabilities. 
@@
 
-Disabled people may rely on different types of assistive technology in their day to day life. 
-Assistive Technology is any form of technology (e.g. software, hardware) that helps people with disabilities perform certain activities.
+The second learning objective echoes our argument on accessible data representation beyond one-type modality.
+People with disabilities may rely on different types of assistive technology in their day to day life.
+Assistive Technology is any form of technology (e.g., software, hardware) that helps people with disabilities perform certain activities.
 Examples of assistive technology include but not limited to walking sticks, wheel chairs and screen readers.
-In incorporating accessibility to our course contents, we wanted students to be familiar with assistive technologies, and since our focus has been on visual accessibility, it was a natural fit to include screen readers.
-A screen reader is an assistive technology that supports blind or visually impaired people in using their computer by reading out the contents. 
-Screen readers are built-in Windows (Settings > Ease of Access > Narrator, and then turn on the toggle under Use Narrator) and Mac OS (System Preferences > Accessibility > Voice Over > Enable Voice Over) and can be used by students.
-
-Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency [@kimAccessibleVisualizationDesign2021]. 
-Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and sometimes people with learning disabilities [@kimAccessibleVisualizationDesign2021; @marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
-We argue that instructors need to teach multiple data representation methods so that all students of all abilities can benefit from multi-sensory data augmentation.
-When shaping our curricular content we have considered multi data representations and their relation to accessibility.
-
-In summary, our curricular goal has been to ensure that the current and the next generation of data scientists, provide public-facing outputs (websites, analysis reports, etc.) that are visually accessible.
-For this goal, we set three specific learning objectives as follows. Students should
-
-1. get familiar with Americans with Disabilities Act;
-2. use at least one assistive technology (i.e. screen reader);
-3. consider different representations of data.
+As we noted above, we introduced screen reader as one of these examples when our students were working on data visualization project; however, we acknowledge that various assistive technologies, such as speech recognition and video captioning, could be integrated into data science courses in diverse contexts (e.g., data collection and data presentation contexts).
 
@@
 The third objective, however, is data science specific.
-In the rest of the section we focus on teaching different representations of data further.
+In the remaining section of this paper, we will further focus on the third objective by illustrating how we have taught multiple data representations.
 
@@
 Images without alt text markup are not readily accessible to assistive technologies (i.e., screen readers; refreshable braille displays; text-based voice commands) and web search parsers. 
-However, appropriately designed alt text can not only give assistive technology users minimum access to visualized data, but also provide all students with a much richer context<!--jy will cite-->.
+However, appropriately designed alt text can not only give assistive technology users minimum access to visualized data, but also provide all students with a much richer context [@lundgardAccessibleVisualizationNatural2022].
 
 There are two ways to create alt text: (1) auto-generated alt text; and (2) manual alt text.
-
 Currently, `BrailleR::VI()` [@R-BrailleR] function has a capability to auto-generate alt text for basic R graphics and ggplot objects (see Figure \ref{fig:VI-example}).

### 2022-07-28 manuscript.Rmd
clarify accessibility
clean up figures

@@
 
-Despite the recent recommendations, accessibility has not yet found its place in data science course materials.
+Despite the recent recommendations, accessibility from disabilities perspectives has not yet found its place in data science course materials.
 One main reason for this could be that accessibility is often covered in the computer sciences [@kawas2019teaching] or the digital arts [@barata2019inclusion] if at all covered.
@@
 
-```{r VI-example, fig.cap = "Automatic data verbalization.", fig.alt = "Bar chart of penguins species with its verbal description automatically generated by `BrailleR::VI()` function.", echo=TRUE, include=TRUE, warning=FALSE, cache=TRUE}
+```{r VI-example, fig.cap = "Sample histogram for automatic data verbalization.", fig.alt = "Bar chart of penguins species with its verbal description automatically generated by `BrailleR::VI()` function.", echo=TRUE, include=TRUE, warning=FALSE, cache=TRUE, fig.height=3}
 library(ggplot2)
@@
 
-```{r plot-xy, fig.cap="Visualizing the relationship of two numerical variables with scatter plot.", fig.alt="Scatter plot representing the positive relationship between x and y variables.", echo=FALSE}
+```{r plot-xy, fig.cap="Visualizing the relationship of two numerical variables with scatter plot.", fig.alt="Scatter plot representing the positive relationship between x and y variables.", echo=FALSE, fig.height=3}
 # Prepare variables

### 2022-07-28 manuscript.Rmd
clean up more figures

@@
 
-```{r plot-xy, fig.cap="Visualizing the relationship of two numerical variables with scatter plot.", fig.alt="Scatter plot representing the positive relationship between x and y variables.", echo=FALSE, fig.height=3}
+```{r plot-xy, fig.cap="Visualizing the relationship of two numerical variables with scatter plot.", fig.alt="Scatter plot representing the positive relationship between x and y variables.", echo=FALSE, fig.height=5}
 # Prepare variables
@@
 
-```{r tactile-producing-code, echo=TRUE, include=TRUE, cache=TRUE}
+<!--added eval=FALSE to the chunk below because the output and messages were irrelevant-->
+
+```{r tactile-producing-code, echo=TRUE, include=TRUE, cache=TRUE, eval=FALSE}
+
 library(palmerpenguins)
@@
 
-brl_begin("figure/tactile-boxplot.pdf")
-with(
-  penguins,
-  boxplot(body_mass_g ~ species)
-)
+brl_begin("tactile-boxplot.pdf")
+with(penguins, boxplot(body_mass_g ~ species))
 brl_end()
@@
 
-```{r tactile-example, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="tactile-producing-code", fig.cap="Ready-to-print tactile graph in braille.", fig.alt="Boxplot showing the different body mass between penguins species"}
+```{r echo=FALSE}
+library(tactileR)
+```
+
+```{r tactile-example, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="tactile-producing-code", fig.cap="Ready-to-print tactile graph in braille.", fig.alt="Boxplot showing the different body mass between penguins species", , fig.align = "center", out.height="50%"}
 knitr::include_graphics("figure/tactile-boxplot.pdf")

### 2022-07-28 manuscript.Rmd
clean up tactualization and sonification

@@
 Our definition of sonification refers to a data representation method using stereo sound and various audible patterns.
-For example, Figure \ref{fig:plot-xy} can be made audible by using `sonify` package [@R-sonify] like below (audio file is available at \url{\audio{}}).
+For example, Figure \ref{fig:plot-xy} can be made audible by using `sonify` package [@R-sonify] like below.
+The output is an audio which can also be accesses as an  audio file  at \url{\audio{}}.
 
@@
 Although this requires some expensive embossing hardware (e.g., braille embosser or Swell Form Heating Machine), this may be one of the most effective and accessible data representations for those who are blind or visually impaired.
-One of the authors in this paper, developed a simple R package, called `tactileR` [@R-tactileR] that can generate a ready-to-emboss pdf file from R graphics and ggplot objects. 
+The `tactileR` [@R-tactileR] package in R can generate a ready-to-emboss pdf file from R graphics and ggplot objects. 
 Students and instructors can produce their own tactile graphics by using the following functions: `tactileR::brl_begin()`; [their graphic object]; `tactileR::brl_end()` (see Figure \ref{fig:tactile-example}).
+For those who do not have access to embossing hardware, the learning process on tactualization can be supported with a video that shows the process of printing a tactile graph available at \url{\video{}}.
 
-<!--added eval=FALSE to the chunk below because the output and messages were irrelevant-->
-
-```{r tactile-producing-code, echo=TRUE, include=TRUE, cache=TRUE, eval=FALSE}
+```{r tactile-producing-code, echo=TRUE, include=TRUE, cache=TRUE, message=FALSE, results='hide'}
 
@@
 brl_end()
-```
-
-```{r echo=FALSE}
-library(tactileR)
 ```

### 2022-07-28 manuscript.Rmd
Closes #4

@@
 Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and people with learning disabilities.
-We argue that instructors need to teach multiple data representation methods so that all students of all abilities can benefit from multi-sensory data augmentation.
+We argue that instructors need to teach multiple data representation methods so that all students can produce data products that are more accessible.
 In this paper, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
@@
 Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and sometimes people with learning disabilities [@kimAccessibleVisualizationDesign2021; @marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
-We argue that instructors need to teach multiple data representation methods so that all students of all abilities can benefit from multi-sensory data augmentation.
+We argue that instructors need to teach multiple data representation methods so that all students can produce data products that are more accessible. 
 To this end, we have considered multi data representations and their relation to accessibility in our courses, and that's what we will demonstrate throughout this paper.

### 2022-07-29 manuscript.Rmd
Closes #7

@@
 
-```{r penguins-basic, echo=FALSE, fig.cap="Visualization showing the relationship between flipper length (mm) and bill length (mm) of penguins for three different species of penguins", fig.alt="A scatterplot showing the relationship between flipper length in mm on the x axis and bill length in mm on the y axis.
+```{r penguins-basic, echo=FALSE, fig.cap="Sample scatterplot showing the relationship between flipper length (mm) and bill length (mm) of penguins for three different species of penguins", fig.alt="Sample scatterplot showing the relationship between flipper length in mm on the x axis and bill length in mm on the y axis.
 Flipper lengths vary from about 170 to 230 and bill lengths vary from about 35 to 60.
@@
 The scatterplots are the same as the scatterplot in the previous figure except for colors.
-From left to right and top to bottom the plots read deutanomly, protanomly, tritanomly, and desaturated", warning=FALSE, echo=TRUE}
+From left to right and top to bottom the plots read deutanomly, protanomly, tritanomly, and desaturated".
+The deutanomly plot shows two sets of points that are close to brownish green and one set of points that are violet.
+The protanomly plot is similar in colors, one of the brownish green is closer to gray.
+In the tritanomly plot, it is difficult to distinguish the green/blue colors.
+There is also a pink color that stands out.
+In the desaturated plot all the points are gray and the different species are indistinguishable., warning=FALSE, echo=TRUE}
 colorblindr::cvd_grid(fig)
@@
 
-For other types of graphics that  `BrailleR` doesn’t support, we ask our students to manually create an alt text.
-We rely on Amy Cesal's “simple formula for writing alt text for data visualization” [@cesal] , which is as follows: “alt= _Chart type_ of _type of data_ where _reason for including chart_.”
-She also provides a great example based on this formula: “Bar chart of gun murders per 100,000 people where America’s murder rate is 6 times worse than Canada, and 30 times Australia.” We highly recommend other instructors try out this alt text formula with their students for manual text description.
+Even though, automated alternate text has some use with its limitations, it does not often directly convey the message a visualization is displaying.
+Thus, it is important to write alternate text manually and teach students how to write one.
+We rely on Cesal's “simple formula for writing alt text for data visualization” [-@cesal] , which is as follows: “alt= _Chart type_ of _type of data_ where _reason for including chart_.” 
+She also recommends inclusion of a link to data somewhere within text. 
+Similarly, for a meaningful alternate text, @canelon suggest that the description conveys the meaning the data, variables are included on the axes, scale and type of plot are desribed.
 
-<!-- @mine -- Please add example figure with with manual alt text. -->
+Given these recommendations, we provide the alternate text we wrote for Figure  \ref{fig:penguins-basic}.
+"Sample scatterplot showing the relationship between flipper length in mm on the x axis and bill length in mm on the y axis.
+Flipper lengths vary from about 170 to 230 and bill lengths vary from about 35 to 60.
+Overall there is a moderate positive relationship.
+Each data point is colored differently for three species as Adelie, Chindtrap and Gentoo.
+Adelie have short flipper length and short bill length in comparison.
+Chinstrap have low flipper length but high bill length.
+Gentoo have high flipper length and high bill length.
+Data points are in red, green, and blue for Adelie, Chinstrap and Gentoo penguins respectively."
+Note that we specify colors as they would be relevant as readers switch from Figure  \ref{fig:penguins-basic} to Figure  \ref{fig:penguins-okabe}.
 
@@
 
-```{r plot-xy, fig.cap="Visualizing the relationship of two numerical variables with scatter plot.", fig.alt="Scatter plot representing the positive relationship between x and y variables.", echo=FALSE, fig.height=5}
+```{r plot-xy, fig.cap="Visualizing the relationship of two numerical variables with scatter plot.", fig.alt="Scatter plot representing the positive relationship between x and y variables.
+There are five points on the plot at x,y values as 1,1 - 2,2 - 3,3 - 4,4 - 5,5", echo=FALSE, fig.height=5}
 # Prepare variables
@@
 
-```{r tactile-example, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="tactile-producing-code", fig.cap="Ready-to-print tactile graph in braille.", fig.alt="Boxplot showing the different body mass between penguins species", , fig.align = "center", out.height="50%"}
+```{r tactile-example, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="tactile-producing-code", fig.cap="Ready-to-print tactile graph in braille.", fig.alt="Boxplot showing the different body mass between penguins species.
+The plot has three boxes for each Penguin species.
+On the x and y-axes, labels and values are displayed in braille.", , fig.align = "center", out.height="50%"}
 knitr::include_graphics("figure/tactile-boxplot.pdf")

### 2022-07-29 manuscript.Rmd
fix citation ??? issue

@@
 bibliography: ["bib/references.bib", "bib/packages.bib"]
+csl: "bib/chicago-author-date.csl"
 output:
   rticles::jds_article: 
+    citation_package: default
     includes:
@@
     base_format: rticles::jds_article # for using bookdown features like \@ref()
+    citation_package: default
 ---
@@
 
-```{r penguins-cvd, echo=FALSE, fig.cap="Color Blindness Simulation with colorblindr", fig.alt="A 2 by 2 grid of four scatterplots.
+```{r penguins-cvd, fig.cap="Color Blindness Simulation with colorblindr", fig.alt="A 2 by 2 grid of four scatterplots.
 The scatterplots are the same as the scatterplot in the previous figure except for colors.
-From left to right and top to bottom the plots read deutanomly, protanomly, tritanomly, and desaturated".
+From left to right and top to bottom the plots read deutanomly, protanomly, tritanomly, and desaturated.
 The deutanomly plot shows two sets of points that are close to brownish green and one set of points that are violet.
@@
 There is also a pink color that stands out.
-In the desaturated plot all the points are gray and the different species are indistinguishable., warning=FALSE, echo=TRUE}
+In the desaturated plot all the points are gray and the different species are indistinguishable.", warning=FALSE, echo=TRUE}
 colorblindr::cvd_grid(fig)

### 2022-07-30 manuscript.Rmd
clean whitespace
update abstract from talk to manuscript

@@
 In this paper, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
-As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we will share specific examples as well as open-access resources that can be utilized by other data science instructors.
+As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we share specific examples that can be utilized by other data science instructors.
 keywords: 
@@
 ```{r tactile-producing-code, echo=TRUE, include=TRUE, cache=TRUE, message=FALSE, results='hide'}
-
 library(palmerpenguins)

### 2022-07-30 manuscript.Rmd
add a paragraph on job readiness in intro

@@
 However, a consensus on what should be included in such curricula has not yet been reached. 
-In the latest version of *Computing Competencies for Undergraduate Data Science Curricula* by ACM Data Science Task Force, accessibility is a recommended content for the data science curricula [@danyluk2021computing].
+One set of guidelines is written by the ACM Data Science Task Force is the  *Computing Competencies for Undergraduate Data Science Curricula*.
+In this report, accessibility is a "foundational consideration" at the knowledge level when it comes to displaying data [@danyluk2021computing].
+
+Training on accessibility is also an important job-readiness aspect of curriculum design. 
+The Partnership on Employment & Accessible Technology (PEAT) states in their Accessible Technology Skills Gap Report [-@peat] that 84% of industry correspondents that they work with say "it is important or very important for them to hire developers and designers with accessible technology skills".
+Even though this statistic maybe different specifically for data scientists, the data science field is broad with intersection in careers as developers (e.g.
+data dashboard developers) and designers (e.g.
+data visualization experts).
+The PEAT also state that 60% of industry respondents "it was difficult or very difficult for their organization to find job candidates with accessibility skills".
 

### 2022-07-31 manuscript.Rmd
fix tactile production path
change title

@@
 title: |
-  Teaching Visual Accessibility in the Introductory Data Science Classes: Why, What, When, and How
+  Teaching Visual Accessibility in Introductory Data Science Classes with Multi-Modal Data Representations
 shorttitle: Teaching Accessibility in Data Science # nolint # nolint
@@
 
-brl_begin("tactile-boxplot.pdf")
+brl_begin("figure/tactile-boxplot.pdf")
 with(penguins, boxplot(body_mass_g ~ species))

### 2022-07-31 manuscript.Rmd
fix street address

@@
 - id: 1
-  institution: "School of Information Sciences, University of Illinois Urbana-Champaign, 614 E. Daniel St. Urbana-Champaign, IL, USA"
+  institution: "School of Information Sciences, University of Illinois Urbana-Champaign, 614 E. Daniel St., Champaign, IL 61820, USA"
 - id: 2

### 2022-07-31 manuscript.Rmd
Grammarly audit first-round

@@
   Teaching Visual Accessibility in Introductory Data Science Classes with Multi-Modal Data Representations
-shorttitle: Teaching Accessibility in Data Science # nolint # nolint
+shorttitle: Teaching Accessibility in Data Science
 author:
@@
   Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency.
-Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and people with learning disabilities.
+Such vision-dependent output may cause critical barriers against those who are blind and visually impaired and people with learning disabilities.
 We argue that instructors need to teach multiple data representation methods so that all students can produce data products that are more accessible.
-In this paper, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
+In this paper, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not, they can have foundational exposure to accessibility.
 As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we share specific examples that can be utilized by other data science instructors.
@@
 However, a consensus on what should be included in such curricula has not yet been reached. 
-One set of guidelines is written by the ACM Data Science Task Force is the  *Computing Competencies for Undergraduate Data Science Curricula*.
+One of the few guidelines written by the ACM Data Science Task Force is the *Computing Competencies for Undergraduate Data Science Curricula*.
 In this report, accessibility is a "foundational consideration" at the knowledge level when it comes to displaying data [@danyluk2021computing].
@@
 Training on accessibility is also an important job-readiness aspect of curriculum design. 
-The Partnership on Employment & Accessible Technology (PEAT) states in their Accessible Technology Skills Gap Report [-@peat] that 84% of industry correspondents that they work with say "it is important or very important for them to hire developers and designers with accessible technology skills".
-Even though this statistic maybe different specifically for data scientists, the data science field is broad with intersection in careers as developers (e.g.
-data dashboard developers) and designers (e.g.
-data visualization experts).
-The PEAT also state that 60% of industry respondents "it was difficult or very difficult for their organization to find job candidates with accessibility skills".
+The Partnership on Employment & Accessible Technology (PEAT) states in their Accessible Technology Skills Gap Report [-@peat] that 84% of industry correspondents that they work with say "it is important or very important for them to hire developers and designers with accessible technology skills."
+Even though this statistic may be different specifically for data scientists, data science field is broad with the intersection in careers as developers (e.g., data dashboard developers) and designers (e.g., data visualization experts).
+The PEAT also states that 60% of industry respondents "it was difficult or very difficult for their organization to find job candidates with accessibility skills."
 
-Despite the recent recommendations, accessibility from disabilities perspectives has not yet found its place in data science course materials.
-One main reason for this could be that accessibility is often covered in the computer sciences [@kawas2019teaching] or the digital arts [@barata2019inclusion] if at all covered.
-However, data science is an interdisciplinary field that is taught in various disciplines including computer science, statistics, business schools, political science, and biological sciences [@schwab2021data].
-Thus, we believe that the resources and conversations around accessibility have not yet extended to the broader data science education community at a scale that it should.
+Despite the recent recommendations, accessibility from disabilities perspectives has not yet found its place in the data science course materials.
+One main reason for this could be that accessibility is often covered in the computer sciences [@kawas2019teaching] or the digital arts [@barata2019inclusion], if at all covered.
+However, data science is an interdisciplinary field that is taught in various disciplines, including computer science, statistics, business schools, political science, and biological sciences [@schwab2021data].
+Thus, we believe that the resources and conversations around accessibility have not yet extended to the broader data science education community at the scale that it should.
 For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on accessibility of data.
 
-In this manuscript, we argue that accessibility should be taught early in the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
+In this manuscript, we argue that accessibility should be taught early in the data science curriculum so that regardless of whether learners major in data science or not, they can have foundational exposure to accessibility.
 As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we share specific examples that can be utilized by other data science instructors.
-We teach R [@R-base] as the main language in our courses thus our examples are shaped around using R, however, instructors using a different set of language(s) or even teaching language-agnostic courses may still find the content applicable for their own courses.
+We teach R [@R-base] as the main language in our courses; thus, our examples are shaped around using R.
+However, instructors using a different set of language(s) or even teaching language-agnostic courses may still find the content applicable for their own courses.
 
@@
 Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency [@kimAccessibleVisualizationDesign2021]. 
-Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and sometimes people with learning disabilities [@kimAccessibleVisualizationDesign2021; @marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
+Such vision-dependent output may cause critical barriers against blind and visually impaired people and sometimes people with learning disabilities [@kimAccessibleVisualizationDesign2021; @marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
 We argue that instructors need to teach multiple data representation methods so that all students can produce data products that are more accessible. 
-To this end, we have considered multi data representations and their relation to accessibility in our courses, and that's what we will demonstrate throughout this paper.
+To this end, we have considered multi-data representations and their relation to accessibility in our courses, and that's what we will demonstrate throughout this paper.
 
@@
 Screen reader is an assistive technology that supports blind or visually impaired people in using their computer by reading aloud the contents.
-Recently, screen reader is provided as one of the built-in accessibility features on operating systems, such as Windows Narrator and Apple's VoiceOver.
+Recently, screen reader has been provided as one of the built-in accessibility features on operating systems, such as Windows Narrator and Apple's VoiceOver.
 
-However, we also want to make our teaching as inclusive as possible to other types of dis/abilities so that we are not just confined with issues of visual disabilities.
+However, we also want to make our teaching as inclusive as possible to other types of dis/abilities so that we are not just confined to issues of visual disabilities.
 
-Therefore, our curricular goal is to broadly ensure that the current and the next generation of data scientists, provide public-facing outputs, including websites and analysis reports, in accessible forms and representations for people with diverse abilities.
+Therefore, our curricular goal is to broadly ensure that the current and the next generation of data scientists provide public-facing outputs, including websites and analysis reports, in accessible forms and representations for people with diverse abilities.
 To achieve this goal, we set three specific and minimal learning objectives like below.
@@
 
-The first objective provides our students with a chance to not just learn legislative concepts, but empathy building concerning diversity issue.
+The first objective provides our students with a chance to learn legislative concepts and empathy building concerning diversity issues.
 As both the authors teach data science in the United States, it was a natural choice for us to make students aware of the Americans with Disabilities Act (ADA).
-The ADA is a civil rights law that prohibits discrimination of individuals with disabilities. 
+The ADA is a civil rights law that prohibits discrimination against individuals with disabilities. 
 Regardless of their own disability status, as future data scientists, we expect our students to produce data science products that are accessible to all, not only because it is the right thing to do but also because they may have future responsibilities as data scientists to comply to ADA.
@@
 The second learning objective echoes our argument on accessible data representation beyond one-type modality.
-People with disabilities may rely on different types of assistive technology in their day to day life.
+People with disabilities may rely on different types of assistive technology in their day-to-day life.
 Assistive Technology is any form of technology (e.g., software, hardware) that helps people with disabilities perform certain activities.
-Examples of assistive technology include but not limited to walking sticks, wheel chairs and screen readers.
-As we noted above, we introduced screen reader as one of these examples when our students were working on data visualization project; however, we acknowledge that various assistive technologies, such as speech recognition and video captioning, could be integrated into data science courses in diverse contexts (e.g., data collection and data presentation contexts).
+Examples of assistive technology include, but not limited to, walking sticks, wheelchairs, and screen readers.
+As we noted above, we introduced screen reader as one of these examples when our students were working on a data visualization project; however, we acknowledge that various assistive technologies, such as speech recognition and video captioning, could be integrated into data science courses in diverse contexts (e.g., data collection and data presentation contexts).
 
-It is important to note that the first two objectives are not data science specific but are related to accessibility in the broader scale. 
+It is important to note that the first two objectives are not data science specific but are related to accessibility on a broader scale. 
 The third objective, however, is data science specific.
@@
 Despite their use by many, data visualizations are not accessible to all. 
-Colors are an important part of data visualizations and an aspect that can limit accessibility for color blind people. 
-When teaching data visualizations we include three basic rules for incorporating colors into data visualization. Students should be able to
+Colors are an important part of data visualizations and an aspect that can limit accessibility for color-blind people. 
+When teaching data visualizations, we include three basic rules for incorporating colors into data visualization. Students should be able to
 
@@
 
-The `colorblindr` package [@R-colorblindr] provides a set of functions that can be useful in teaching and practising color blindness accessibility. 
+The `colorblindr` package [@R-colorblindr] provides a set of functions that can be useful in teaching and practicing color blindness accessibility. 
 For instance, `colorblindr::cvd_grid()` function provides color-deficiency simulations of a given plot. 
-In the code chunk below the plot is named as `fig`. 
-The code creates a grid of four plots where each plot shows a different color vision deficiency including deutanomly, protanomly, tritanomly, and desaturated.
-This grid allows those without any color vision deficiencies experience what data visualizations would look like to a sighted person that has color blindness.
+In the code chunk below, the plot is named as `fig`. 
+The code creates a grid of four plots where each plot shows a different color vision deficiency, including deuteranomaly, protanomaly, tritanopia, and desaturated.
+This grid allows those without any color vision deficiencies to experience what data visualizations would look like to a sighted person that has color blindness.
 
@@
 
-For their data visualizations, instead of relying on the default color selection of R packages, students are introduced to the Okabe-Ito color palette which is known to be accessible to people with color vision deficiencies [@okabe].
+For their data visualizations, instead of relying on the default color selection of R packages, students are introduced to the Okabe-Ito color palette, which is known to be accessible to people with color vision deficiencies [@okabe].
 One can find specific colors in the pallette by the R code `palette.colors(palette = "Okabe-Ito")`.
 In addition, students can be introduced to `colorblindr::scale_color_OkabeIto()` and `colorblindr::scale_fill_OkabeIto()` to automatically switch to using this color pallette by adding a layer to a ggplot presentation.
-For instance, the plot in Figure \ref{fig:penguins-basic} can be switched to Okabe-Ito color pallette as shown in Figure \ref{fig:penguins-okabe} by using the code below.
+For instance, the plot in Figure \ref{fig:penguins-basic} can be switched to the Okabe-Ito color palette as shown in Figure \ref{fig:penguins-okabe} by using the code below.
 
@@
 
-Last but not least, students should learn not to rely only on color to differentiate data points, in this case Species. 
-For instance, they can be introduced to shapes and have data points that are circle, triangle, square for each species. 
+Last but not least, students should learn not to rely only on color to differentiate data points, in this case, Species. 
+For instance, they can be introduced to shapes and have data points that are circles, triangles, and squares for each species. 
 This can also be extended to line graphs as having continuous or dashed lines.
@@
 Those who teach interactive data visualizations may also consider covering cognitive and motor accessibility. 
-Our focus here is on visual accessibility but it is worth noting that all subdomains of accessibility are relevant to data science learners and can be incorporated into the curriculum as seen fit.
+Our focus here is on visual accessibility, but it is worth noting that all subdomains of accessibility are relevant to data science learners and can be incorporated into the curriculum as seen fit.
 
@@
 Images without alt text markup are not readily accessible to assistive technologies (i.e., screen readers; refreshable braille displays; text-based voice commands) and web search parsers. 
-However, appropriately designed alt text can not only give assistive technology users minimum access to visualized data, but also provide all students with a much richer context [@lundgardAccessibleVisualizationNatural2022].
+However, appropriately designed alt text can not only give assistive technology users minimum access to visualized data but also provide all students with a much richer context [@lundgardAccessibleVisualizationNatural2022].
 
@@
 
-Even though, automated alternate text has some use with its limitations, it does not often directly convey the message a visualization is displaying.
+Even though the automated alternate text has some use with its limitations, it does not often directly convey the message a visualization is displaying.
 Thus, it is important to write alternate text manually and teach students how to write one.
-We rely on Cesal's “simple formula for writing alt text for data visualization” [-@cesal] , which is as follows: “alt= _Chart type_ of _type of data_ where _reason for including chart_.” 
-She also recommends inclusion of a link to data somewhere within text. 
-Similarly, for a meaningful alternate text, @canelon suggest that the description conveys the meaning the data, variables are included on the axes, scale and type of plot are desribed.
+We rely on Cesal's “simple formula for writing alt text for data visualization” [-@cesal], which is as follows: “alt= _Chart type_ of _type of data_ where _reason for including chart_.” 
+She also recommends the inclusion of a link to data somewhere within the text. 
+Similarly, for a meaningful alternate text, @canelon suggest that the description should convey the meaning of the data, variables included on the axes, scale, and type of plot.
 
 Given these recommendations, we provide the alternate text we wrote for Figure  \ref{fig:penguins-basic}.
-"Sample scatterplot showing the relationship between flipper length in mm on the x axis and bill length in mm on the y axis.
-Flipper lengths vary from about 170 to 230 and bill lengths vary from about 35 to 60.
+"Sample scatterplot showing the relationship between flipper length in mm on the x-axis and bill length in mm on the y-axis.
+Flipper lengths vary from about 170 to 230, and bill lengths vary from about 35 to 60.
 Overall there is a moderate positive relationship.
-Each data point is colored differently for three species as Adelie, Chindtrap and Gentoo.
-Adelie have short flipper length and short bill length in comparison.
-Chinstrap have low flipper length but high bill length.
-Gentoo have high flipper length and high bill length.
-Data points are in red, green, and blue for Adelie, Chinstrap and Gentoo penguins respectively."
+Each data point is colored differently for three species as Adelie, Chinstrap, and Gentoo.
+Adelie has a short flipper length and short bill length in comparison.
+Chinstrap has low flipper length but high bill length.
+Gentoo has a high flipper length and high bill length.
+Data points are in red, green, and blue for Adelie, Chinstrap, and Gentoo penguins, respectively."
 Note that we specify colors as they would be relevant as readers switch from Figure  \ref{fig:penguins-basic} to Figure  \ref{fig:penguins-okabe}.
@@
 For example, Figure \ref{fig:plot-xy} can be made audible by using `sonify` package [@R-sonify] like below.
-The output is an audio which can also be accesses as an  audio file  at \url{\audio{}}.
+The output is audio which can also be accessed as an audio file at \url{\audio{}}.
 
@@
 
-In this example, a scatter plot is represented by a sound. Position on the X axis is communicated using the left-to-right stereo effect; the pitch of the sound indicates position on the Y axis.
+In this example, a scatter plot is represented by sound. Position on the X axis is communicated using the left-to-right stereo effect; the pitch of the sound indicates the position on the Y axis.
 
@@
 For instance, instructors incorporating accessibility should modify assessment instructions and rubrics to include accessibility. 
-This will prevent students just hearing about accessibility in the lecture then forgetting about it. 
+This will prevent students from just hearing about accessibility in the lecture and then forgetting about it. 
 Ideally, the assessments should not treat accessibility as a learning objective at a single point in the academic term. 
@@
 
-In this manuscript we have focused on domain-specific (i.e. data science) knowledge of accessibility using a particular language (R). 
+In this manuscript, we have focused on domain-specific (i.e., data science) knowledge of accessibility using a particular language (R). 
 Depending on the courses taught, instructors may want to extend the accessibility content to non-data contexts. 
-In addition they may adopt the content in their courses with other languages such as Python and SAS or even in their language-agnostic courses. 
+In addition, they may adopt the content in their courses with other languages such as Python and SAS or even in their language-agnostic courses. 
 Overall, if taught well, a student learning accessibility content, for instance, should be able to include alternate text in their PowerPoint presentation in their Anthropology class. 
@@
 
-Although, we have mainly focused on tools and different representations of data, in order to have students develop accessible content, a more crucial accessibility feature is reproducible workflows. 
+Although we have mainly focused on tools and different representations of data to have students develop accessible content, a more crucial accessibility feature is reproducible workflows. 
 Scholars often focus on reproducibility and replication of scientific findings. 
-We believe that reproducibility can often serve as an accessibility feature in data science context. 
-A reproducible workflow that is open-source provides an opportunity for users who are blind or have visual impairment an opportunity to examine data in their preferred form of data representations.
-Thus in addition to teaching different representations of data, instructors may consider teaching reproducibility as well.
+We believe reproducibility can often serve as an accessibility feature in a data science context. 
+A reproducible open-source workflow allows users who are blind or have a visual impairment to examine data in their preferred form of data representation.
+Thus, in addition to teaching different representations of data, instructors may also consider teaching reproducibility.
 
@@
 
-We firmly believe that an accessible future in the field of data science requires educators to deliberately teach accessibility as part of the curriculum. 
+We believe that an accessible future in data science requires educators to teach accessibility as part of the curriculum deliberately. 
 We also acknowledge that what we have presented here is only a small portion of the intersection of visual accessibility and data science. 

### 2022-07-31 manuscript.Rmd
final review

@@
 The Partnership on Employment & Accessible Technology (PEAT) states in their Accessible Technology Skills Gap Report [-@peat] that 84% of industry correspondents that they work with say "it is important or very important for them to hire developers and designers with accessible technology skills."
-Even though this statistic may be different specifically for data scientists, data science field is broad with the intersection in careers as developers (e.g., data dashboard developers) and designers (e.g., data visualization experts).
+Even though this statistic may be different specifically for data scientists, the data science field is broad with the intersection in careers as developers (e.g., data dashboard developers) and designers (e.g., data visualization experts).
 The PEAT also states that 60% of industry respondents "it was difficult or very difficult for their organization to find job candidates with accessibility skills."
@@
 However, data science is an interdisciplinary field that is taught in various disciplines, including computer science, statistics, business schools, political science, and biological sciences [@schwab2021data].
-Thus, we believe that the resources and conversations around accessibility have not yet extended to the broader data science education community at the scale that it should.
-For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on accessibility of data.
+Thus, we believe that the resources and conversations around accessibility have not yet extended to the broader data science education community at the scale that they should.
+For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on the accessibility of data.
 
@@
 Assistive Technology is any form of technology (e.g., software, hardware) that helps people with disabilities perform certain activities.
-Examples of assistive technology include, but not limited to, walking sticks, wheelchairs, and screen readers.
+Examples of assistive technology include, but are not limited to, walking sticks, wheelchairs, and screen readers.
 As we noted above, we introduced screen reader as one of these examples when our students were working on a data visualization project; however, we acknowledge that various assistive technologies, such as speech recognition and video captioning, could be integrated into data science courses in diverse contexts (e.g., data collection and data presentation contexts).
@@
 
-Data visualization is the most common representation of data.
-Despite their use by many, data visualizations are not accessible to all. 
+Data visualization is the most common representation of data. Despite their use by many, data visualizations are not accessible to all. 
 Colors are an important part of data visualizations and an aspect that can limit accessibility for color-blind people. 
@@
 Currently, `BrailleR::VI()` [@R-BrailleR] function has a capability to auto-generate alt text for basic R graphics and ggplot objects (see Figure \ref{fig:VI-example}).
-Supported graphic types include as follows: `graphics::hist()`, `graphics::boxplot()`, `ggplot2::geom_bar()`, `ggplot2::geom_histogram()`, and `ggplot2::geom_boxplot()`.
+Some of the supported graphic types include: `graphics::hist()`, `graphics::boxplot()`, `ggplot2::geom_bar()`, `ggplot2::geom_histogram()`, and `ggplot2::geom_boxplot()`.
 
@@
 
-```{r tactile-producing-code, echo=TRUE, include=TRUE, cache=TRUE, message=FALSE, results='hide'}
+```{r tactile-producing-code, echo=TRUE, include=TRUE, cache=TRUE, message=FALSE, results="hide"}
 library(palmerpenguins)
@@
 Educators interested in teaching accessibility can follow the organization's work at https://teachaccess.org/. 
-The source code of this manuscript can be found on a GitHub repository at \url{\repo{}}.
+The reproducible R source code of this manuscript can be found on a GitHub repository at \url{\repo{}}.
 

### 2022-07-31 header.tex
Closes #8
clean up tactualization and sonification
some rewording and language edit

@@
 % For anonymity
-\newcommand{\repo}[1]{https://github.com/mdogucu/teaching-accessibility-manuscript}
+\newcommand{\repo}[1]{https://github.com/jooyoungseo/teaching-accessibility-manuscript}
+\newcommand{\audio}[1]{https://jooyoungseo.github.io/teaching-accessibility-manuscript/audio/sonify-example.wav}
+\newcommand{\video}[1]{https://www.youtube.com/watch?v=ClI555l4Z1M&ab_channel=JooYoungSeo}
 

### 2022-07-31 manuscript.Rmd
some rewording and language edit

@@
 We argue that instructors need to teach multiple data representation methods so that all students can produce data products that are more accessible.
-In this paper, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not, they can have foundational exposure to accessibility.
+In this paper, we argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not, they can have foundational exposure to accessibility.
 As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we share specific examples that can be utilized by other data science instructors.
@@
 However, a consensus on what should be included in such curricula has not yet been reached. 
-One of the few guidelines written by the ACM Data Science Task Force is the *Computing Competencies for Undergraduate Data Science Curricula*.
+One of the few guidelines on the topic is written by the ACM Data Science Task Force and titled *Computing Competencies for Undergraduate Data Science Curricula*.
 In this report, accessibility is a "foundational consideration" at the knowledge level when it comes to displaying data [@danyluk2021computing].
@@
 The Partnership on Employment & Accessible Technology (PEAT) states in their Accessible Technology Skills Gap Report [-@peat] that 84% of industry correspondents that they work with say "it is important or very important for them to hire developers and designers with accessible technology skills."
-Even though this statistic may be different specifically for data scientists, the data science field is broad with the intersection in careers as developers (e.g., data dashboard developers) and designers (e.g., data visualization experts).
+Even though this statistic may be different specifically for data scientists, the data science field is broad with the intersection in careers as developers (e.g., data dashboard and package developers) and designers (e.g., data visualization experts).
 The PEAT also states that 60% of industry respondents "it was difficult or very difficult for their organization to find job candidates with accessibility skills."
@@
 Thus, we believe that the resources and conversations around accessibility have not yet extended to the broader data science education community at the scale that they should.
-For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on the accessibility of data.
+For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on the accessing data.
 
@@
 The third objective, however, is data science specific.
-In the remaining section of this paper, we will further focus on the third objective by illustrating how we have taught multiple data representations.
+In the remaining section of this paper, we will focus further on the third objective by illustrating how we have taught multiple data representations.
 
@@
 The code creates a grid of four plots where each plot shows a different color vision deficiency, including deuteranomaly, protanomaly, tritanopia, and desaturated.
-This grid allows those without any color vision deficiencies to experience what data visualizations would look like to a sighted person that has color blindness.
+This grid allows those without any color vision deficiencies to experience what data visualizations would look like to a sighted person that has color defiency.
 
@@
 She also recommends the inclusion of a link to data somewhere within the text. 
-Similarly, for a meaningful alternate text, @canelon suggest that the description should convey the meaning of the data, variables included on the axes, scale, and type of plot.
+Similarly, for a meaningful alternate text, @canelon suggest that the description should convey the meaning of the data, variables on the axes, scale, and type of plot should be included.
 

### 2022-08-01 arxiv/manuscript-arxiv.Rmd
add arxiv version with ASA template

--- 2022-07-31 manuscript.Rmd
@@
 ---
-title: |
-  Teaching Visual Accessibility in Introductory Data Science Classes with Multi-Modal Data Representations
-shorttitle: Teaching Accessibility in Data Science
-author:
+title: Teaching Visual Accessibility in Introductory Data Science Classes with Multi-Modal Data Representations
+
+# to produce blinded version set to 1
+blinded: 0
+
+authors: 
 - name: JooYoung Seo 
-  affiliation: "1"  # Must be quoted when including comma.
-  email: "jseo1005@illinois.edu"  # Only corresponding authors require email.
+  affiliation: School of Information Sciences, University of Illinois Urbana-Champaign
+
 - name: Mine Dogucu
-  affiliation: "2,3"
+  affiliation: |
+    | Department of Statistical Science, University College London 
+    | Department of Statistics, University of California Irvine
 
-affiliation:
-- id: 1
-  institution: "School of Information Sciences, University of Illinois Urbana-Champaign, 614 E. Daniel St., Champaign, IL 61820, USA"
-- id: 2
-  institution: "Department of Statistical Science, University College London"
-- id: 3
-  institution: "Department of Statistics, University of California Irvine"
+keywords:
+  - Data representations
+  - Curriculum
+  - R
 
@@
 As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we share specific examples that can be utilized by other data science instructors.
-keywords: 
-  - Data representations
-  - Curriculum
-  - R
-
-# the following edits should be done by Journal typesetters
-page: 1  # set the first page number
-month: July
-year: 2020
-volume: xx
-issue: xx
-doi: "xx.xxxx/xxxxxxxxx"
-received: "April, 2020"  # optional: comment it out if no received date
-accepted: "May, 2020"  # optional: comment it out if no accepted date
 
 bibliography: ["bib/references.bib", "bib/packages.bib"]
-csl: "bib/chicago-author-date.csl"
 output:
-  rticles::jds_article: 
-    citation_package: default
-    includes:
-      in_header: header.tex #header-hidden.tex can be used when creating an anonymized version for review
   bookdown::pdf_book:
-    base_format: rticles::jds_article # for using bookdown features like \@ref()
-    citation_package: default
+    base_format: rticles::asa_article
+    extra_dependencies: ["graphics", "float"]
 ---
@@
 Training on accessibility is also an important job-readiness aspect of curriculum design.
-The Partnership on Employment & Accessible Technology (PEAT) states in their Accessible Technology Skills Gap Report [-@peat] that 84% of industry correspondents that they work with say "it is important or very important for them to hire developers and designers with accessible technology skills."
+The Partnership on Employment \& Accessible Technology (PEAT) states in their Accessible Technology Skills Gap Report [-@peat] that 84% of industry correspondents that they work with say "it is important or very important for them to hire developers and designers with accessible technology skills."
 Even though this statistic may be different specifically for data scientists, the data science field is broad with the intersection in careers as developers (e.g., data dashboard and package developers) and designers (e.g., data visualization experts).
@@
 For example, Figure \ref{fig:plot-xy} can be made audible by using `sonify` package [@R-sonify] like below.
-The output is audio which can also be accessed as an audio file at \url{\audio{}}.
+The output is audio which can also be accessed as an audio file at \url{https://jooyoungseo.github.io/teaching-accessibility-manuscript/audio/sonify-example.wav}.
 
@@
 Students and instructors can produce their own tactile graphics by using the following functions: `tactileR::brl_begin()`; [their graphic object]; `tactileR::brl_end()` (see Figure \ref{fig:tactile-example}).
-For those who do not have access to embossing hardware, the learning process on tactualization can be supported with a video that shows the process of printing a tactile graph available at \url{\video{}}.
+For those who do not have access to embossing hardware, the learning process on tactualization can be supported with a video that shows the process of printing a tactile graph available at \url{https://www.youtube.com/watch?v=ClI555l4Z1M&ab_channel=JooYoungSeo}.
 
@@
 Educators interested in teaching accessibility can follow the organization's work at https://teachaccess.org/. 
-The reproducible R source code of this manuscript can be found on a GitHub repository at \url{\repo{}}.
+The reproducible R source code of this manuscript can be found on a GitHub repository at \url{https://github.com/jooyoungseo/teaching-accessibility-manuscript}.
 

### 2022-08-04 arxiv/manuscript-arxiv/manuscript-arxiv.tex
change bib to bbl for arxiv submission

--- 2022-08-01 arxiv/manuscript-arxiv.Rmd
@@
----
-title: Teaching Visual Accessibility in Introductory Data Science Classes with Multi-Modal Data Representations
+% !TeX program = pdfLaTeX
+\documentclass[12pt]{article}
+\usepackage{amsmath}
+\usepackage{graphicx,psfrag,epsf}
+\usepackage{enumerate}
+\usepackage{natbib}
+\usepackage{textcomp}
+\usepackage[hyphens]{url} % not crucial - just used below for the URL
+\usepackage{hyperref}
 
-# to produce blinded version set to 1
-blinded: 0
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{0}
 
-authors: 
-- name: JooYoung Seo 
-  affiliation: School of Information Sciences, University of Illinois Urbana-Champaign
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.3in}%
+\addtolength{\topmargin}{-.8in}%
 
-- name: Mine Dogucu
-  affiliation: |
-    | Department of Statistical Science, University College London 
-    | Department of Statistics, University of California Irvine
+%% load any required packages here
 
-keywords:
-  - Data representations
-  - Curriculum
-  - R
+% Pandoc syntax highlighting
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 
-abstract: |
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% From pandoc table feature
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+
+\usepackage{graphics}
+\usepackage{float}
+
+\begin{document}
+
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\if0\blind
+{
+  \title{\bf Teaching Visual Accessibility in Introductory Data Science Classes with Multi-Modal Data Representations}
+
+  \author{
+        JooYoung Seo \\
+    School of Information Sciences, University of Illinois Urbana-Champaign\\
+     and \\     Mine Dogucu \\
+    Department of Statistical Science, University College London\\
+Department of Statistics, University of California Irvine\\
+      }
+  \maketitle
+} \fi
+
+\if1\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf Teaching Visual Accessibility in Introductory Data Science Classes with Multi-Modal Data Representations}
+  \end{center}
+  \medskip
+} \fi
+
+\bigskip
+\begin{abstract}
 Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency.
@@
 As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we share specific examples that can be utilized by other data science instructors.
+\end{abstract}
 
-bibliography: ["bib/references.bib", "bib/packages.bib"]
-output:
-  bookdown::pdf_book:
-    base_format: rticles::asa_article
-    extra_dependencies: ["graphics", "float"]
----
+\noindent%
+{\it Keywords:} Data representations, Curriculum, R
+\vfill
 
-```{r setup, include=FALSE, echo=FALSE, purl=FALSE}
-knitr::opts_chunk$set(echo = FALSE, fig.path = "figure/")
+\newpage
+\spacingset{1.45} % DON'T change the spacing!
 
-# deactivate addition of preamble when this format is used with bookdown
-options(bookdown.theorem.preamble = FALSE)
-```
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-```{r pkg, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, purl=FALSE}
-# Load required packages
-## Any packages loaded in this chunk will automatically be added to `bib/packages.bib` for citation use
-library(ggplot2)
-library(scales)
-library(BrailleR)
-library(sonify)
-library(tactileR)
-library(tidyverse)
-library(shiny)
-library(knitr)
-library(gt)
-library(colorblindr)
-library(palmerpenguins)
-```
-
-# Introduction
-
-According to LinkedIn's U.S. Emerging Jobs Report, data scientists rank among the top emerging jobs [-@linkedin].
+According to LinkedIn's U.S. Emerging Jobs Report, data scientists rank among the top emerging jobs \citeyearpar{linkedin}.
 With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula.
 However, a consensus on what should be included in such curricula has not yet been reached.
-One of the few guidelines on the topic is written by the ACM Data Science Task Force and titled *Computing Competencies for Undergraduate Data Science Curricula*.
-In this report, accessibility is a "foundational consideration" at the knowledge level when it comes to displaying data [@danyluk2021computing].
+One of the few guidelines on the topic is written by the ACM Data Science Task Force and titled \emph{Computing Competencies for Undergraduate Data Science Curricula}.
+In this report, accessibility is a ``foundational consideration'' at the knowledge level when it comes to displaying data \citep{danyluk2021computing}.
 
 Training on accessibility is also an important job-readiness aspect of curriculum design.
-The Partnership on Employment \& Accessible Technology (PEAT) states in their Accessible Technology Skills Gap Report [-@peat] that 84% of industry correspondents that they work with say "it is important or very important for them to hire developers and designers with accessible technology skills."
+The Partnership on Employment \& Accessible Technology (PEAT) states in their Accessible Technology Skills Gap Report \citeyearpar{peat} that 84\% of industry correspondents that they work with say ``it is important or very important for them to hire developers and designers with accessible technology skills.''
 Even though this statistic may be different specifically for data scientists, the data science field is broad with the intersection in careers as developers (e.g., data dashboard and package developers) and designers (e.g., data visualization experts).
-The PEAT also states that 60% of industry respondents "it was difficult or very difficult for their organization to find job candidates with accessibility skills."
+The PEAT also states that 60\% of industry respondents ``it was difficult or very difficult for their organization to find job candidates with accessibility skills.''
 
 Despite the recent recommendations, accessibility from disabilities perspectives has not yet found its place in the data science course materials.
-One main reason for this could be that accessibility is often covered in the computer sciences [@kawas2019teaching] or the digital arts [@barata2019inclusion], if at all covered.
-However, data science is an interdisciplinary field that is taught in various disciplines, including computer science, statistics, business schools, political science, and biological sciences [@schwab2021data].
+One main reason for this could be that accessibility is often covered in the computer sciences \citep{kawas2019teaching} or the digital arts \citep{barata2019inclusion}, if at all covered.
+However, data science is an interdisciplinary field that is taught in various disciplines, including computer science, statistics, business schools, political science, and biological sciences \citep{schwab2021data}.
 Thus, we believe that the resources and conversations around accessibility have not yet extended to the broader data science education community at the scale that they should.
-For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on the accessing data.
+For instance, many introductory data science courses \citep{yan2019first, baumer2015data}, textbooks \citep{adhikari2019computational, wickham2016r}, or curricula \citep{schwab2021data} either fail to mention accessibility at all or only focus on the accessing data.
 
@@
 As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we share specific examples that can be utilized by other data science instructors.
-We teach R [@R-base] as the main language in our courses; thus, our examples are shaped around using R.
+We teach R \citep{R-base} as the main language in our courses; thus, our examples are shaped around using R.
 However, instructors using a different set of language(s) or even teaching language-agnostic courses may still find the content applicable for their own courses.
 
-# Learning Goals and Objectives
+\hypertarget{learning-goals-and-objectives}{%
+\section{Learning Goals and Objectives}\label{learning-goals-and-objectives}}
 
-Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency [@kimAccessibleVisualizationDesign2021]. 
-Such vision-dependent output may cause critical barriers against blind and visually impaired people and sometimes people with learning disabilities [@kimAccessibleVisualizationDesign2021; @marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
+Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency \citep{kimAccessibleVisualizationDesign2021}.
+Such vision-dependent output may cause critical barriers against blind and visually impaired people and sometimes people with learning disabilities \citep{kimAccessibleVisualizationDesign2021, marriottInclusiveDataVisualization2021, leeReachingBroaderAudiences2020}.
 We argue that instructors need to teach multiple data representation methods so that all students can produce data products that are more accessible.
@@
 
-1. get familiar with Americans with Disabilities Act;
-2. use at least one assistive technology;
-3. consider different representations of data.
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  get familiar with Americans with Disabilities Act;
+\item
+  use at least one assistive technology;
+\item
+  consider different representations of data.
+\end{enumerate}
 
@@
 
-## Data Visualization
+\hypertarget{data-visualization}{%
+\subsection{Data Visualization}\label{data-visualization}}
 
@@
 
-1. simulate color blindness (if they are sighted and not color blind);
-2. pick color blind friendly colors;
-3. not only rely on color for differentiating data.
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  simulate color blindness (if they are sighted and not color blind);
+\item
+  pick color blind friendly colors;
+\item
+  not only rely on color for differentiating data.
+\end{enumerate}
 
-We show an example of incorporating these three basic rules by using the Palmer Penguins data [@R-palmerpenguins]. 
+We show an example of incorporating these three basic rules by using the Palmer Penguins data \citep{R-palmerpenguins}.
 We visualize the relationship between flipper length and bill length of penguins and color the data points differently for each species in Figure \ref{fig:penguins-basic}.
 
-```{r penguins-basic, echo=FALSE, fig.cap="Sample scatterplot showing the relationship between flipper length (mm) and bill length (mm) of penguins for three different species of penguins", fig.alt="Sample scatterplot showing the relationship between flipper length in mm on the x axis and bill length in mm on the y axis.
-Flipper lengths vary from about 170 to 230 and bill lengths vary from about 35 to 60.
-Overall there is a moderate positive relationship.
-Each data point is colored differently for three species as Adelie, Chindtrap and Gentoo.
-Adelie have short flipper length and short bill length in comparison.
-Chinstrap have low flipper length but high bill length.
-Gentoo have high flipper length and high bill length.
-Data points are in red, green, and blue for Adelie, Chinstrap and Gentoo penguins respectively.", warning=FALSE, fig.height=3}
-library(ggplot2) # It seems like the pkg chunk is not loading packages. I did not want to make changes there since I do not know what your intention was.
-library(palmerpenguins)
+\begin{figure}
+\includegraphics{figure/penguins-basic-1} \caption{Sample scatterplot showing the relationship between flipper length (mm) and bill length (mm) of penguins for three different species of penguins}\label{fig:penguins-basic}
+\end{figure}
 
-fig <- ggplot(penguins, aes(flipper_length_mm, bill_length_mm, color = species)) +
-  geom_point() +
-  labs(
-    x = "Flipper Length(mm)",
-    y = "Bill Length (mm)",
-    color = "Species"
-  ) +
-  theme_bw()
-
-fig
-```
-
-The `colorblindr` package [@R-colorblindr] provides a set of functions that can be useful in teaching and practicing color blindness accessibility. 
-For instance, `colorblindr::cvd_grid()` function provides color-deficiency simulations of a given plot. 
-In the code chunk below, the plot is named as `fig`. 
+The \texttt{colorblindr} package \citep{R-colorblindr} provides a set of functions that can be useful in teaching and practicing color blindness accessibility.
+For instance, \texttt{colorblindr::cvd\_grid()} function provides color-deficiency simulations of a given plot.
+In the code chunk below, the plot is named as \texttt{fig}.
 The code creates a grid of four plots where each plot shows a different color vision deficiency, including deuteranomaly, protanomaly, tritanopia, and desaturated.
@@
 
-```{r penguins-cvd, fig.cap="Color Blindness Simulation with colorblindr", fig.alt="A 2 by 2 grid of four scatterplots.
-The scatterplots are the same as the scatterplot in the previous figure except for colors.
-From left to right and top to bottom the plots read deutanomly, protanomly, tritanomly, and desaturated.
-The deutanomly plot shows two sets of points that are close to brownish green and one set of points that are violet.
-The protanomly plot is similar in colors, one of the brownish green is closer to gray.
-In the tritanomly plot, it is difficult to distinguish the green/blue colors.
-There is also a pink color that stands out.
-In the desaturated plot all the points are gray and the different species are indistinguishable.", warning=FALSE, echo=TRUE}
-colorblindr::cvd_grid(fig)
-```
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{colorblindr}\SpecialCharTok{::}\FunctionTok{cvd\_grid}\NormalTok{(fig)}
+\end{Highlighting}
+\end{Shaded}
 
-For their data visualizations, instead of relying on the default color selection of R packages, students are introduced to the Okabe-Ito color palette, which is known to be accessible to people with color vision deficiencies [@okabe].
-One can find specific colors in the pallette by the R code `palette.colors(palette = "Okabe-Ito")`.
-In addition, students can be introduced to `colorblindr::scale_color_OkabeIto()` and `colorblindr::scale_fill_OkabeIto()` to automatically switch to using this color pallette by adding a layer to a ggplot presentation.
+\begin{figure}
+\includegraphics{figure/penguins-cvd-1} \caption{Color Blindness Simulation with colorblindr}\label{fig:penguins-cvd}
+\end{figure}
+
+For their data visualizations, instead of relying on the default color selection of R packages, students are introduced to the Okabe-Ito color palette, which is known to be accessible to people with color vision deficiencies \citep{okabe}.
+One can find specific colors in the pallette by the R code \texttt{palette.colors(palette\ =\ "Okabe-Ito")}.
+In addition, students can be introduced to \texttt{colorblindr::scale\_color\_OkabeIto()} and \texttt{colorblindr::scale\_fill\_OkabeIto()} to automatically switch to using this color pallette by adding a layer to a ggplot presentation.
 For instance, the plot in Figure \ref{fig:penguins-basic} can be switched to the Okabe-Ito color palette as shown in Figure \ref{fig:penguins-okabe} by using the code below.
 
-```{r penguins-okabe, fig.cap="Scatterplot Using Okabe-Ito Color Pallette", fig.alt="The scatterplots are the same as the scatterplot in the original penguins plot except for colors.
-Data points are in orange, sky blue, and bluish green for Adelie, Chinstrap and Gentoo penguins respectively.", warning=FALSE, fig.height=3, echo=TRUE}
-fig + colorblindr::scale_color_OkabeIto()
-```
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{fig }\SpecialCharTok{+}\NormalTok{ colorblindr}\SpecialCharTok{::}\FunctionTok{scale\_color\_OkabeIto}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+\includegraphics{figure/penguins-okabe-1} \caption{Scatterplot Using Okabe-Ito Color Pallette}\label{fig:penguins-okabe}
+\end{figure}
 
@@
 
-## Data verbalization (i.e., alt text)
+\hypertarget{data-verbalization-i.e.-alt-text}{%
+\subsection{Data verbalization (i.e., alt text)}\label{data-verbalization-i.e.-alt-text}}
 
-Our definition of "data verbalization" refers to a way of representing data patterns using verbal description with alternative text (i.e., alt text). 
+Our definition of ``data verbalization'' refers to a way of representing data patterns using verbal description with alternative text (i.e., alt text).
 Images without alt text markup are not readily accessible to assistive technologies (i.e., screen readers; refreshable braille displays; text-based voice commands) and web search parsers.
-However, appropriately designed alt text can not only give assistive technology users minimum access to visualized data but also provide all students with a much richer context [@lundgardAccessibleVisualizationNatural2022].
+However, appropriately designed alt text can not only give assistive technology users minimum access to visualized data but also provide all students with a much richer context \citep{lundgardAccessibleVisualizationNatural2022}.
 
 There are two ways to create alt text: (1) auto-generated alt text; and (2) manual alt text.
-Currently, `BrailleR::VI()` [@R-BrailleR] function has a capability to auto-generate alt text for basic R graphics and ggplot objects (see Figure \ref{fig:VI-example}).
-Some of the supported graphic types include: `graphics::hist()`, `graphics::boxplot()`, `ggplot2::geom_bar()`, `ggplot2::geom_histogram()`, and `ggplot2::geom_boxplot()`.
+Currently, \texttt{BrailleR::VI()} \citep{R-BrailleR} function has a capability to auto-generate alt text for basic R graphics and ggplot objects (see Figure \ref{fig:VI-example}).
+Some of the supported graphic types include: \texttt{graphics::hist()}, \texttt{graphics::boxplot()}, \texttt{ggplot2::geom\_bar()}, \texttt{ggplot2::geom\_histogram()}, and \texttt{ggplot2::geom\_boxplot()}.
 
-```{r VI-example, fig.cap = "Sample histogram for automatic data verbalization.", fig.alt = "Bar chart of penguins species with its verbal description automatically generated by `BrailleR::VI()` function.", echo=TRUE, include=TRUE, warning=FALSE, cache=TRUE, fig.height=3}
-library(ggplot2)
-library(BrailleR)
-library(palmerpenguins)
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(ggplot2)}
+\FunctionTok{library}\NormalTok{(BrailleR)}
+\FunctionTok{library}\NormalTok{(palmerpenguins)}
 
-# Create graph
-g_species <- ggplot(data = penguins, mapping = aes(x = species)) +
-  geom_bar()
+\CommentTok{\# Create graph}
+\NormalTok{g\_species }\OtherTok{\textless{}{-}} \FunctionTok{ggplot}\NormalTok{(}\AttributeTok{data =}\NormalTok{ penguins, }\AttributeTok{mapping =} \FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ species)) }\SpecialCharTok{+}
+  \FunctionTok{geom\_bar}\NormalTok{()}
 
-g_species
+\NormalTok{g\_species}
+\end{Highlighting}
+\end{Shaded}
 
-# Obtain auto verbalization
-# (Added feature in BrailleR >= 0.32.1): `BrailleR::VI()` function is automatically called against supported graphics if BrailleR package is currently loaded in R session
-# BrailleR::VI(g_species)
-```
+\begin{figure}
+\includegraphics{figure/VI-example-1} \caption{Sample histogram for automatic data verbalization.}\label{fig:VI-example}
+\end{figure}
+
+\begin{verbatim}
+## This is an untitled chart with no subtitle or caption.
+## It has x-axis 'species' with labels Adelie, Chinstrap and Gentoo.
+## It has y-axis 'count' with labels 0, 50, 100 and 150.
+## The chart is a bar chart with 3 vertical bars.
+## Bar 1 is centered horizontally at Adelie, and spans vertically from 0 to 152.
+## Bar 2 is centered horizontally at Chinstrap, and spans vertically from 0 to 68.
+## Bar 3 is centered horizontally at Gentoo, and spans vertically from 0 to 124.
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# Obtain auto verbalization}
+\CommentTok{\# (Added feature in BrailleR \textgreater{}= 0.32.1): \textasciigrave{}BrailleR::VI()\textasciigrave{} function is automatically called against supported graphics if BrailleR package is currently loaded in R session}
+\CommentTok{\# BrailleR::VI(g\_species)}
+\end{Highlighting}
+\end{Shaded}
 
@@
 Thus, it is important to write alternate text manually and teach students how to write one.
-We rely on Cesal's “simple formula for writing alt text for data visualization” [-@cesal], which is as follows: “alt= _Chart type_ of _type of data_ where _reason for including chart_.” 
+We rely on Cesal's ``simple formula for writing alt text for data visualization'' \citeyearpar{cesal}, which is as follows: ``alt= \emph{Chart type} of \emph{type of data} where \emph{reason for including chart}.''
 She also recommends the inclusion of a link to data somewhere within the text.
-Similarly, for a meaningful alternate text, @canelon suggest that the description should convey the meaning of the data, variables on the axes, scale, and type of plot should be included.
+Similarly, for a meaningful alternate text, \citet{canelon} suggest that the description should convey the meaning of the data, variables on the axes, scale, and type of plot should be included.
 
 Given these recommendations, we provide the alternate text we wrote for Figure \ref{fig:penguins-basic}.
-"Sample scatterplot showing the relationship between flipper length in mm on the x-axis and bill length in mm on the y-axis.
+``Sample scatterplot showing the relationship between flipper length in mm on the x-axis and bill length in mm on the y-axis.
 Flipper lengths vary from about 170 to 230, and bill lengths vary from about 35 to 60.
@@
 Gentoo has a high flipper length and high bill length.
-Data points are in red, green, and blue for Adelie, Chinstrap, and Gentoo penguins, respectively."
+Data points are in red, green, and blue for Adelie, Chinstrap, and Gentoo penguins, respectively.''
 Note that we specify colors as they would be relevant as readers switch from Figure \ref{fig:penguins-basic} to Figure \ref{fig:penguins-okabe}.
 
-Once alt text is prepared either programmatically or manually, it can be embedded in static and dynamic plot objects in R, such as `ggplot2::labs(..., alt = "[alt text]")`, 
-`shiny::renderPlot(..., alt = "[alt tex]")` [@R-shiny].
-We also ask our students to provide alt text within their R Markdown chunks by using `fig.alt = "[alt text]"` `knitr` chunk option [@knitr2014], and within Pandoc-flavor/Common Markdown with `![alt text here.](path/to/file)` syntax [@macfarlanePandoc2022].
+Once alt text is prepared either programmatically or manually, it can be embedded in static and dynamic plot objects in R, such as \texttt{ggplot2::labs(...,\ alt\ =\ "{[}alt\ text{]}")},
+\texttt{shiny::renderPlot(...,\ alt\ =\ "{[}alt\ tex{]}")} \citep{R-shiny}.
+We also ask our students to provide alt text within their R Markdown chunks by using \texttt{fig.alt\ =\ "{[}alt\ text{]}"} \texttt{knitr} chunk option \citep{knitr2014}, and within Pandoc-flavor/Common Markdown with \texttt{!{[}alt\ text\ here.{]}(path/to/file)} syntax \citep{macfarlanePandoc2022}.
 
-## Data sonification using sound
+\hypertarget{data-sonification-using-sound}{%
+\subsection{Data sonification using sound}\label{data-sonification-using-sound}}
 
 Our definition of sonification refers to a data representation method using stereo sound and various audible patterns.
-For example, Figure \ref{fig:plot-xy} can be made audible by using `sonify` package [@R-sonify] like below.
+For example, Figure \ref{fig:plot-xy} can be made audible by using \texttt{sonify} package \citep{R-sonify} like below.
 The output is audio which can also be accessed as an audio file at \url{https://jooyoungseo.github.io/teaching-accessibility-manuscript/audio/sonify-example.wav}.
 
-```{r sonify-x-y, echo=TRUE, include=TRUE, cache=TRUE}
-# Representing the relationship between the two numerical variables via stereo sound
-library(sonify())
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# Representing the relationship between the two numerical variables via stereo sound}
+\FunctionTok{library}\NormalTok{(}\FunctionTok{sonify}\NormalTok{())}
 
-# Prepare variables
-x <- 1:5
-y <- 1:5
+\CommentTok{\# Prepare variables}
+\NormalTok{x }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{5}
+\NormalTok{y }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{5}
 
-sonify(x, y)
-```
+\FunctionTok{sonify}\NormalTok{(x, y)}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-```{r plot-xy, fig.cap="Visualizing the relationship of two numerical variables with scatter plot.", fig.alt="Scatter plot representing the positive relationship between x and y variables.
-There are five points on the plot at x,y values as 1,1 - 2,2 - 3,3 - 4,4 - 5,5", echo=FALSE, fig.height=5}
-# Prepare variables
-x <- 1:5
-y <- 1:5
+\begin{figure}
+\includegraphics{figure/plot-xy-1} \caption{Visualizing the relationship of two numerical variables with scatter plot.}\label{fig:plot-xy}
+\end{figure}
 
-# Representing the relationship between x and y variables via plot visualization
-plot(x, y)
-```
-
-## Data tactualization using Swell Form machine
+\hypertarget{data-tactualization-using-swell-form-machine}{%
+\subsection{Data tactualization using Swell Form machine}\label{data-tactualization-using-swell-form-machine}}
 
@@
 Although this requires some expensive embossing hardware (e.g., braille embosser or Swell Form Heating Machine), this may be one of the most effective and accessible data representations for those who are blind or visually impaired.
-The `tactileR` [@R-tactileR] package in R can generate a ready-to-emboss pdf file from R graphics and ggplot objects. 
-Students and instructors can produce their own tactile graphics by using the following functions: `tactileR::brl_begin()`; [their graphic object]; `tactileR::brl_end()` (see Figure \ref{fig:tactile-example}).
+The \texttt{tactileR} \citep{R-tactileR} package in R can generate a ready-to-emboss pdf file from R graphics and ggplot objects.
+Students and instructors can produce their own tactile graphics by using the following functions: \texttt{tactileR::brl\_begin()}; {[}their graphic object{]}; \texttt{tactileR::brl\_end()} (see Figure \ref{fig:tactile-example}).
 For those who do not have access to embossing hardware, the learning process on tactualization can be supported with a video that shows the process of printing a tactile graph available at \url{https://www.youtube.com/watch?v=ClI555l4Z1M&ab_channel=JooYoungSeo}.
 
-```{r tactile-producing-code, echo=TRUE, include=TRUE, cache=TRUE, message=FALSE, results="hide"}
-library(palmerpenguins)
-library(tactileR)
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(palmerpenguins)}
+\FunctionTok{library}\NormalTok{(tactileR)}
 
-brl_begin("figure/tactile-boxplot.pdf")
-with(penguins, boxplot(body_mass_g ~ species))
-brl_end()
-```
+\FunctionTok{brl\_begin}\NormalTok{(}\StringTok{"figure/tactile{-}boxplot.pdf"}\NormalTok{)}
+\FunctionTok{with}\NormalTok{(penguins, }\FunctionTok{boxplot}\NormalTok{(body\_mass\_g }\SpecialCharTok{\textasciitilde{}}\NormalTok{ species))}
+\FunctionTok{brl\_end}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
 
-```{r tactile-example, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="tactile-producing-code", fig.cap="Ready-to-print tactile graph in braille.", fig.alt="Boxplot showing the different body mass between penguins species.
-The plot has three boxes for each Penguin species.
-On the x and y-axes, labels and values are displayed in braille.", , fig.align = "center", out.height="50%"}
-knitr::include_graphics("figure/tactile-boxplot.pdf")
-```
+\begin{figure}
 
-# Delivery of Content
+{\centering \includegraphics[height=0.5\textheight]{figure/tactile-boxplot}
+
+}
+
+\caption{Ready-to-print tactile graph in braille.}\label{fig:tactile-example}
+\end{figure}
+
+\hypertarget{delivery-of-content}{%
+\section{Delivery of Content}\label{delivery-of-content}}
 
@@
 
-# Closing Remarks
+\hypertarget{closing-remarks}{%
+\section{Closing Remarks}\label{closing-remarks}}
 
@@
 The second co-author was supported through a Teach Access grant to develop curricular materials on teaching accessibility as part of the data science curriculum.
-Educators interested in teaching accessibility can follow the organization's work at https://teachaccess.org/. 
+Educators interested in teaching accessibility can follow the organization's work at \url{https://teachaccess.org/}.
 The reproducible R source code of this manuscript can be found on a GitHub repository at \url{https://github.com/jooyoungseo/teaching-accessibility-manuscript}.
 
-# References {-}
+\bibliographystyle{agsm}
+\bibliography{bib/references.bib,bib/packages.bib}
 
-```{r pkg_bib, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="pkg"}
-knitr::write_bib(file = "bib/packages.bib")
-```
+\end{document}

### 2022-10-19 revision_letter.Rmd
add reviewer comments to the revision letter template

---
title       : "Teaching Visual Accessibility in Introductory Data Science Classes with Multi-Modal Data Representations"
authors     : "JooYoung Seo & Mine Dogucu"
journal     : "Journal of Data Science"
manuscript  : "JDS2207-022"

class       : "draft"
output      : papaja::revision_letter_pdf
bibliography: ["bib/packages.bib", "bib/references.bib"]
---

Dear Dr. Claire McKay Bowen,

thank you very much for taking the time to consider our manuscript for publication at _`r rmarkdown::metadata$journal`_.
In the following we address your and each reviewers' concerns point-by-point.

# AE

\RC{
    The authors preface their paper with the statement that "In this paper, we argue that accessibility should be taught as early as the introductory course as part of the data science curriculum." This argument is never presented. There is no discussion as to WHY this topic should be included in an introductory course. Why can't it be part of an upper level course?
    }

This is our response

\Assignment{First Author}
\WorkInProgress
\Easy

\RC{
    The authors close with the powerful statement "We believe that an accessible future in data science requires educators to teach accessibility as part of the curriculum deliberately"--but they have failed to present what incorporating it deliberately entails.
    }

This is our response.

> This is a section quoted from the revised manuscript to illustrate the change.

\Assignment{Second Author}
\AlmostDone
\Medium

\RC{
    I believe there are two courses of action the authors could take with this paper: a) develop and incorporate a significant argument for accessibility as part of a data science curriculum b) reframe this paper as an experiment in incorporating accessibility methods in an intro course As it currently stands, it falls somewhere in between and falls short of providing a meaningful presentation of either.
    }

This is our response

\Assignment{First Author}
\Done
\TimeConsuming
\Hard
\NotEstimated

# Reviewer

\RC{
    What are the significances of teaching visual accessibility at the 'introductory' data science level (as opposed to introducing them at a higher level of data science courses after students have a solid foundation in the discipline?
}

This is our response

\Assignment{First Author}
\WorkInProgress
\Easy

\RC{
If the visual aid accessibility was introduced in the 'introductory' courses, will students see it again in higher-level courses? In other words, would other faculty members who teach higher levels continue to use this in their courses? Students would not retain the information if they were only to see visual aid accessibility at the beginning of their data science journey.
}

This is our response

\Assignment{First Author}
\WorkInProgress
\Easy

\RC{
As visual aid accessibility is an essential part of various disciplines and career paths (science, engineering, business, arts and letters, etc.), would it be more beneficial to have a separate course (as a university-core or university-elective course) teaching visual aid accessibility rather than to integrate it into the foundational data science courses?
}

This is our response

\Assignment{First Author}
\WorkInProgress
\Easy

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::

### 2022-11-03 revision_letter.Rmd
capitalize
more ideas

@@
 
-thank you very much for taking the time to consider our manuscript for publication at _`r rmarkdown::metadata$journal`_.
+Thank you very much for taking the time to consider our manuscript for publication at _`r rmarkdown::metadata$journal`_.
 In the following we address your and each reviewers' concerns point-by-point.
@@
 
-This is our response.
+This is our response: For example, instructors can include accessibility as one of their evaluation criteria for course assignments or make it as a chance to obtain bonus credits for the students' visualization work.
 
@@
 
-This is our response
+This is our response: need to create a section dededicated to a11y, and say that we are presenting this paper as a demonstration how we integrated a11y into our courses.
 
@@
 
-This is our response
+This is our response: good point so it is critical to share the same principle across different and with upper level courses .
 

### 2022-11-04 manuscript.Rmd
add more bib library

@@
 
-bibliography: ["bib/references.bib", "bib/packages.bib"]
+bibliography: ["bib/references.bib", "bib/packages.bib", "bib/teachinga11y.bib"]
 csl: "bib/chicago-author-date.csl"

### 2022-11-04 manuscript.Rmd
add annotated bibliography

@@
 ```
+
+# Lit Review
+
+## @shinoharaWhoTeachesAccessibility2018
+
+### RP
+
+Industry demand for software developers with knowledge of accessibility has increased substantially in recent years.
+However, there is little knowledge about the prevalence of higher education teaching about accessibility or faculty's perceived barriers to teaching accessibility.
+
+### RQ
+
+### DC
+
+we surveyed 14,176 computing and information science faculty in the United States.
+We received a representative sample of at least one response from 318 of the 352 institutions we surveyed, totaling 1,857 responses.
+
+### Findings
+
+175 institutions (50%) had at least one instructor teaching accessibility and that no fewer than 2.5% of faculty overall teach accessibility.
+Faculty that teach accessibility are twice as likely to be female, to have expertise in HCI and software engineering, and to know people with disabilities.
+
+The most critical barriers to teaching accessibility that faculty reported were the absence of clear and discipline-specific accessibility learning objectives and the lack of faculty knowledge about accessibility.
+
+Faculty desired resources that were specific to the areas of computing in which they teach rather than general accessibility resources and guidelines.

### 2022-11-04 manuscript.Rmd
add more bib
Annotated bib

@@
 Faculty desired resources that were specific to the areas of computing in which they teach rather than general accessibility resources and guidelines.
+
+## @kawas2019teaching
+
+### RP
+
+Most CS students learn little about accessibility in higher education; this is partly because most CS faculty know little about accessibility.
+
+higher education CS faculty lack a model of professional development for learning to teach new topics.
+
+### DC
+
+conducted 18 semi-structured interviews with U.S. CS faculty, asking them to explore a prototype of a web-based professional development tool that linked accessibility topics to CS topics.
+
+we investigated the feasibility of a "micro" professional development model for teaching accessibility in CS courses that could be used at scale.
+
+### Findings
+
+found that many organizational factors limited faculty's autonomy to integrate accessibility in many of their courses.
+
+also found that individual values and knowledge constrained faculty's ability and willingness to both learn and integrate accessibility topics into their courses.
+
+ However, many faculty expressed desire to teach accessibility in their courses if they had access to even basic accessibility content and materials to use in their courses.

### 2022-11-05 manuscript.Rmd
some notes in lit review

@@
 
+* RQ1. Who is teaching accessibility?
+* RQ2. What barriers do faculty see to teaching accessibility?
+* RQ3. What factors predict who is teaching accessibility?
+
 ### DC
 
-we surveyed 14,176 computing and information science faculty in the United States.
+surveyed 14,176 computing and information science faculty in the United States.
 We received a representative sample of at least one response from 318 of the 352 institutions we surveyed, totaling 1,857 responses.
@@
 Faculty desired resources that were specific to the areas of computing in which they teach rather than general accessibility resources and guidelines.
+
+### Quotes
+
+> Consortia like TeachAccess (teachaccess.org), led by Yahoo!, Facebook, Microsoft, Google, and Adobe, represent this increased demand for accessibility knowledge, and are leading efforts to increase the number of computer and information science faculty teaching accessibility.
+
+> In parallel, professional societies such as ABET [abet.org] have recently added accessibility into engineering design accreditation requirements.
+
+> Incorporating instruction on accessibility at scale requires two major elements: (1) methods for how to teach accessibility in computer science (CS) and related fields, and (2) large scale adoption and implementation of these methods by faculty.
+<!-- In conclusion, we have just demonstrated some methods to make data science visualization more accessible, focusing on the second element.
+However, we need more broad effort to make it at a large scale as @shinoharaWhoTeachesAccessibility2018 said.
+-->
 
@@
 
-found that many organizational factors limited faculty's autonomy to integrate accessibility in many of their courses.
+many organizational factors limited faculty's autonomy to integrate accessibility in many of their courses.
 
-also found that individual values and knowledge constrained faculty's ability and willingness to both learn and integrate accessibility topics into their courses.
+individual values and knowledge constrained faculty's ability and willingness to both learn and integrate accessibility topics into their courses.
 
 However, many faculty expressed desire to teach accessibility in their courses if they had access to even basic accessibility content and materials to use in their courses.
+
+## @koAccessComputingPromotesTeaching2016#
+
+### RP
+
+### RQ
+
+### DC
+
+## ### Findings

### 2022-11-05 revision_letter.Rmd
some notes in lit review

@@
 output      : papaja::revision_letter_pdf
-bibliography: ["bib/packages.bib", "bib/references.bib"]
+bibliography: ["bib/packages.bib", "bib/references.bib", "bib/teachinga11y.bib"]
 ---
@@
 
-This is our response
+According to @shinoharaWhoTeachesAccessibility2018,
+
+> The most critical barriers to teaching accessibility that faculty reported were the absence of clear and discipline-specific accessibility learning objectives and the lack of faculty knowledge about accessibility.
+
+> Faculty desired resources that were specific to the areas of computing in which they teach rather than general accessibility resources and guidelines.
 

### 2022-11-05 manuscript.Rmd
more annotation and quotes

@@
 
+> some have incorporated accessibility in software engineering courses [@martin-escalonaIntroductionTopicAccessibility2013; @ludiIntroducingAccessibilityRequirements2007; @ludiAccessEveryoneIntroducing2002], as part of introducing web development [@koAccessComputingPromotesTeaching2016; @rosmaitaAccessibilityFirstNew2006], or in HCI and design courses as a central component of design process [@shinoharaHowDesigningPeople2016; @shinoharaTeachingAccessibilityTechnology2017a].
+
 > Incorporating instruction on accessibility at scale requires two major elements: (1) methods for how to teach accessibility in computer science (CS) and related fields, and (2) large scale adoption and implementation of these methods by faculty.
@@
 -->
+
+> These works suggest that large scale curricular change is not a matter of top-down decree, but one of fostering local change agents who are intimately aware of the culture of specific institutions and who can devise and implement strategies to work within them.
+
+> theories of curricular change argue that change only comes through social and instructional discourse among faculty organically contributing to, and constructing, instructors’ knowledge [@barnettConceptualisingCurriculumChange2001; @kirkTeacherVoiceOwnership2001].
+
+> Kezar and Eckel, in an examination of several institutions, determined that institutional culture could enhance or stymie curriculum changes [@kezarEffectInstitutionalCulture2002].
 

### 2022-11-06 manuscript.Rmd
update annotated bibliography
update annotations

--- 2022-11-05 manuscript.Rmd
@@
 
+Overall, 20% of faculty (375) responded “Yes” to the prompt “Do you teach courses that incorporate topics about accessibility?”
+
 The most critical barriers to teaching accessibility that faculty reported were the absence of clear and discipline-specific accessibility learning objectives and the lack of faculty knowledge about accessibility.
@@
 
+<!-- Increasing interest and importance of accessibility -->
 > Consortia like TeachAccess (teachaccess.org), led by Yahoo!, Facebook, Microsoft, Google, and Adobe, represent this increased demand for accessibility knowledge, and are leading efforts to increase the number of computer and information science faculty teaching accessibility.
@@
 
+<!-- Examples of how accessibility is integrated into computing and information sciences courses -->
+> some have incorporated accessibility in software engineering courses [@martin-escalonaIntroductionTopicAccessibility2013; @ludiIntroducingAccessibilityRequirements2007; @ludiAccessEveryoneIntroducing2002], as part of introducing web development [@koAccessComputingPromotesTeaching2016; @rosmaitaAccessibilityFirstNew2006], or in HCI and design courses as a central component of design process [@shinoharaHowDesigningPeople2016; @shinoharaTeachingAccessibilityTechnology2017a].
+
+<!-- Two efforts are required to make broader impacts for teaching accessibility -->
 > Incorporating instruction on accessibility at scale requires two major elements: (1) methods for how to teach accessibility in computer science (CS) and related fields, and (2) large scale adoption and implementation of these methods by faculty.
@@
 -->
+
+> These works suggest that large scale curricular change is not a matter of top-down decree, but one of fostering local change agents who are intimately aware of the culture of specific institutions and who can devise and implement strategies to work within them.
+
+<!-- Active culture movement is required towards teaching accessibility -->
+> theories of curricular change argue that change only comes through social and instructional discourse among faculty organically contributing to, and constructing, instructors’ knowledge [@barnettConceptualisingCurriculumChange2001; @kirkTeacherVoiceOwnership2001].
+
+> Kezar and Eckel, in an examination of several institutions, determined that institutional culture could enhance or stymie curriculum changes [@kezarEffectInstitutionalCulture2002].
 
@@
 ## ### Findings
+
+## ## @joynerVisualizationAccessibilityWild2022
+
+### rp
+
+### rq
+
+### dc
+
+### ### findings
+
+## @marriottInclusiveDataVisualization2021
+
+### RP
+
+### RQ
+
+### DC
+
+### Findings
+
+### Quotes
+
+<!-- how many people with disabilities in the world? -->
+> A 2011 report from the WHO (https://bit.ly/3raT0kg) estimates that more than 1 billion people—15 percent of the world’s population— live with some form of disability, of whom about one-fifth experience significant difficulties in functioning.
+While it may be obvious that people who are blind or have low vision (BLV) would face challenges in accessing data visualizations, this also holds for other kinds of disability.
+People who have cognitive and learning disabilities (CALD) frequently have trouble making sense of the abstractions and symbolic conventions used in visualizations, and people who lack fine motor coordination or have tremors may be unable to operate interactive visualizations.
+
+<!-- Why is accessible Visualization important? -->
+> this is an equity issue with severe consequences, because a lack of access to visualizations and underlying data impacts health, educational and work opportunities, and lifestyle.
+For instance, the BLV community has identified a lack of access to educational materials [@butlerUnderstandingGraphicalChallenges2017] and online data visualizations about Covid-19 as a significant concern (https://bit.ly/2NJzy0C).
+
+<!-- People are not paying enough attention to accessible viz -->
+> The visualization research community, however, has not paid enough attention to the needs of people with disabilities [@leeReachingBroaderAudiences2020].
+
+<!-- Accessible viz for all beyond disabilities -->
+> Inclusive approaches to visualizations are likely to help more than those people with permanent disabilities.
+There are also hundreds of millions of people with temporary or situational disabilities.
+New, inclusive approaches could readily become “informational curb cuts” that benefit everyone.

### 2022-11-06 manuscript.Rmd
more quotes

@@
 New, inclusive approaches could readily become “informational curb cuts” that benefit everyone.
+
+<!-- Tactile representation for blind people -->
+> The Braille Authority of North America (BANA; http://www.
+brailleauthority.org) recommends the use of tactile graphics to show spatial data.
+These are usually raised line drawings, but more recently, 3D-printed models have been used.
+Research suggests that despite the differences between tactile and visual perception, spatial perception and reasoning is fundamental to the use of both visual and tactile graphics, and that tactile maps, scatter plots, and network diagrams offer similar benefits to the corresponding visual graphic (e.g., @yangTactilePresentationNetwork2020a).
+
+<!-- Tactile production is expensive -->
+> Unfortunately, the cost and time for the production of tactile graphics limits their use for data exploration.
+More recently, researchers have explored the multimodal presentation of graphics on touchscreen devices, providing audio and haptic feedback when the reader’s fingers touch the graphic.
+There is considerable opportunity to explore data-exploration tools leveraging refreshable tactile displays as well as touchscreen-based presentations, as these may provide similar benefits to tactile graphics.

### 2022-11-07 manuscript.Rmd
more bib
put some lit on the back burner

@@
 
-## @koAccessComputingPromotesTeaching2016#
-
-### RP
-
-### RQ
-
-### DC
-
-## ### Findings
-
-## ## @joynerVisualizationAccessibilityWild2022
-
-### rp
-
-### rq
-
-### dc
-
-### ### findings
-
 ## @marriottInclusiveDataVisualization2021
@@
 People who have cognitive and learning disabilities (CALD) frequently have trouble making sense of the abstractions and symbolic conventions used in visualizations, and people who lack fine motor coordination or have tremors may be unable to operate interactive visualizations.
+
+<!-- Three types of major disabilities  affected by inaccessible viz -->
+> There are three main kinds of disability that may affect access to visualizations: visual, cognitive, and motor impairments. It is not uncommon for people to have more than one kind of impairment.
 
@@
 There is considerable opportunity to explore data-exploration tools leveraging refreshable tactile displays as well as touchscreen-based presentations, as these may provide similar benefits to tactile graphics.
+
+<!-- critical disability studies -->
+@mankoffDisabilityStudiesSource2010
+
+<!-- The following is good to review when having more time -->
+
+## ## @joynerVisualizationAccessibilityWild2022
+
+### rp
+
+### rq
+
+### dc
+
+### ### findings
+
+## @koAccessComputingPromotesTeaching2016
+
+### RP
+
+### RQ
+
+### DC
+
+### Findings

### 2022-11-07 manuscript.Rmd
WIP: refine intro section

@@
 
-Training on accessibility is also an important job-readiness aspect of curriculum design.
-The Partnership on Employment & Accessible Technology (PEAT) states in their Accessible Technology Skills Gap Report [-@peat] that 84% of industry correspondents that they work with say "it is important or very important for them to hire developers and designers with accessible technology skills."
-Even though this statistic may be different specifically for data scientists, the data science field is broad with the intersection in careers as developers (e.g., data dashboard and package developers) and designers (e.g., data visualization experts).
+In fact, training on accessibility has been increasingly becoming an important job-readiness aspect in computing and information sciences [@shinoharaWhoTeachesAccessibility2018; @kawasTeachingAccessibilityDesign2019].
+For instance, some notable big tech companies, such as Yahoo!, Facebook, Microsoft, Google, and Adobe, have been leading a TeachAccess (teachaccess.org) Consortia to represent this increased demand for accessibility knowledge and to increase the number of computer and information science faculty teaching accessibility [-@shinoharaWhoTeachesAccessibility2018].
+The Partnership on Employment & Accessible Technology (PEAT), in their Accessible Technology Skills Gap Report [-@peat], also highlights this trend--84% of industry correspondents that they work with say "it is important or very important for them to hire developers and designers with accessible technology skills." Growing interest in and importance of teaching accessibility can be also found in professional societies.
+For example, accessibility has been recently added to engineering design accreditation requirements by Accreditation Board for Engineering and Technology [ABET; abet.org] [@shinoharaWhoTeachesAccessibility2018].
+
+Even though this statistic may be different specifically for data scientists, the data science as a highly interdisciplinary field across computing and information sciences also shares the rising demand for the accessibility preparedness.
+Furthermore, data science is broad with the intersection in careers as developers (e.g., data dashboard and package developers) and designers (e.g., data visualization experts).
+Therefore, the impact of teaching and learning accessibility is as far-reaching as its wide scope.
+
 The PEAT also states that 60% of industry respondents "it was difficult or very difficult for their organization to find job candidates with accessibility skills."
 
-Despite the recent recommendations, accessibility from disabilities perspectives has not yet found its place in the data science course materials.
-One main reason for this could be that accessibility is often covered in the computer sciences [@kawas2019teaching] or the digital arts [@barata2019inclusion], if at all covered.
+<!-- Examples of how accessibility is integrated into computing and information sciences courses -->
+> some have incorporated accessibility in software engineering courses [@martin-escalonaIntroductionTopicAccessibility2013; @ludiIntroducingAccessibilityRequirements2007; @ludiAccessEveryoneIntroducing2002], as part of introducing web development [@koAccessComputingPromotesTeaching2016; @rosmaitaAccessibilityFirstNew2006], or in HCI and design courses as a central component of design process [@shinoharaHowDesigningPeople2016; @shinoharaTeachingAccessibilityTechnology2017a].
+
+Despite the increasing demand for accessibility knowledge in both industries and academia, accessibility from disabilities perspectives has not yet found its place in the data science course materials.
+One main reason for this could be that accessibility is often covered in the computer sciences [@kawasTeachingAccessibilityDesign2019] or the digital arts [@barata2019inclusion], if at all covered.
 However, data science is an interdisciplinary field that is taught in various disciplines, including computer science, statistics, business schools, political science, and biological sciences [@schwab2021data].
@@
 
-## @kawas2019teaching
+## @kawasTeachingAccessibilityDesign2019
 
@@
 ## @marriottInclusiveDataVisualization2021
-
-### RP
-
-### RQ
-
-### DC
-
-### Findings
 

### 2022-11-07 manuscript.Rmd
WIP: update intro

@@
 
-According to LinkedIn's U.S. Emerging Jobs Report, data scientists rank among the top emerging jobs [-@linkedin].
+According to LinkedIn's U.S.
+Emerging Jobs Report, data scientists rank among the top emerging jobs [-@linkedin].
 With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula.
@@
 Therefore, the impact of teaching and learning accessibility is as far-reaching as its wide scope.
+
+Data science consits of iterative process importing, tidying, transforming, visualizing, and modeling data [@wickham2016r].
+While each stage has unique challenges, data visualization may pose fundamental barriers against people with permanent, situational, and temporary disabiliteis (e.g., visual, cognitive, and motor impairments to name a few) in accessing information [@marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
+For example, people with cognitive and learning disabilities often have difficulties understanding the abstractions and symbolic conventions used in visualizations, and people lacking fine motor coordination or having tremors may be unable to operate interactive visualizations not to mention people who are blind and have low vision facing extra challenges in visualization access [@marriottInclusiveDataVisualization2021].
+@marriottInclusiveDataVisualization2021, in their call to action paper, highlights the serious impact of inaccessible visualizations: "this is an equity issue with severe consequences, because a lack of access to visualizations and underlying data impacts health, educational and work opportunities, and lifestyle." <!-- Need to find a specific page number for this quote.
+-->
+
+Despite these
+
+<!-- Accessible viz for all beyond disabilities -->
+> Inclusive approaches to visualizations are likely to help more than those people with permanent disabilities.
+There are also hundreds of millions of people with temporary or situational disabilities.
+New, inclusive approaches could readily become “informational curb cuts” that benefit everyone.
 

### 2022-11-07 manuscript.Rmd
finish intro update

--- 2022-11-07 manuscript.Rmd
@@
 For instance, some notable big tech companies, such as Yahoo!, Facebook, Microsoft, Google, and Adobe, have been leading a TeachAccess (teachaccess.org) Consortia to represent this increased demand for accessibility knowledge and to increase the number of computer and information science faculty teaching accessibility [-@shinoharaWhoTeachesAccessibility2018].
-The Partnership on Employment & Accessible Technology (PEAT), in their Accessible Technology Skills Gap Report [-@peat], also highlights this trend--84% of industry correspondents that they work with say "it is important or very important for them to hire developers and designers with accessible technology skills." Growing interest in and importance of teaching accessibility can be also found in professional societies.
+The Partnership on Employment & Accessible Technology (PEAT), in their Accessible Technology Skills Gap Report [-@peat], also highlights this trend--84% of industry correspondents that they work with say "it is important or very important for them to hire developers and designers with accessible technology skills." The growing interest in and importance of teaching accessibility can also be found in professional education societies.
 For example, accessibility has been recently added to engineering design accreditation requirements by Accreditation Board for Engineering and Technology [ABET; abet.org] [@shinoharaWhoTeachesAccessibility2018].
 
-Even though this statistic may be different specifically for data scientists, the data science as a highly interdisciplinary field across computing and information sciences also shares the rising demand for the accessibility preparedness.
-Furthermore, data science is broad with the intersection in careers as developers (e.g., data dashboard and package developers) and designers (e.g., data visualization experts).
-Therefore, the impact of teaching and learning accessibility is as far-reaching as its wide scope.
+Data science consists of an iterative process of importing, tidying, transforming, visualizing, and modeling data [@wickham2016r].
+While each stage has unique challenges, data visualization may pose fundamental barriers against people with permanent, situational, and temporary disabilities (e.g., visual, cognitive, and motor impairments, to name a few) in accessing information [@marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
+For example, people with cognitive and learning disabilities often have difficulties understanding the abstractions and symbolic conventions used in visualizations, and people lacking fine motor coordination or having tremors may be unable to operate interactive visualizations, not to mention people who are blind and have low vision facing extra challenges in visualization access [@marriottInclusiveDataVisualization2021].
+@marriottInclusiveDataVisualization2021, in their call to action paper, highlights the serious impact of inaccessible visualizations: "This is an equity issue with severe consequences, because a lack of access to visualizations and underlying data impacts health, educational and work opportunities, and lifestyle." <!-- Need to find a specific page number for this quote.
+-->
 
-The PEAT also states that 60% of industry respondents "it was difficult or very difficult for their organization to find job candidates with accessibility skills."
-
-<!-- Examples of how accessibility is integrated into computing and information sciences courses -->
-> some have incorporated accessibility in software engineering courses [@martin-escalonaIntroductionTopicAccessibility2013; @ludiIntroducingAccessibilityRequirements2007; @ludiAccessEveryoneIntroducing2002], as part of introducing web development [@koAccessComputingPromotesTeaching2016; @rosmaitaAccessibilityFirstNew2006], or in HCI and design courses as a central component of design process [@shinoharaHowDesigningPeople2016; @shinoharaTeachingAccessibilityTechnology2017a].
-
-Despite the increasing demand for accessibility knowledge in both industries and academia, accessibility from disabilities perspectives has not yet found its place in the data science course materials.
-One main reason for this could be that accessibility is often covered in the computer sciences [@kawasTeachingAccessibilityDesign2019] or the digital arts [@barata2019inclusion], if at all covered.
-However, data science is an interdisciplinary field that is taught in various disciplines, including computer science, statistics, business schools, political science, and biological sciences [@schwab2021data].
-Thus, we believe that the resources and conversations around accessibility have not yet extended to the broader data science education community at the scale that they should.
+Despite the increasing demand for accessibility knowledge in both industries and academia and the critical impact of accessible visualizations, accessibility has not yet found its place in the data science course and teaching materials.
+One main reason for this could be that accessibility is often incorporated into human-computer interaction (HCI) and design courses [@shinoharaHowDesigningPeople2016; @shinoharaTeachingAccessibilityTechnology2017a], introductory web development [@koAccessComputingPromotesTeaching2016; @rosmaitaAccessibilityFirstNew2006], software engineering courses [@martin-escalonaIntroductionTopicAccessibility2013; @ludiIntroducingAccessibilityRequirements2007; @ludiAccessEveryoneIntroducing2002] or the digital arts [@barata2019inclusion], if at all covered.
+However, data science is an interdisciplinary field that is taught in various disciplines, including not only computer and information sciences but also statistics, business schools, political science, biological sciences, and more [@schwab2021data].
+Thus, we posit that the resources and conversations around accessibility have not yet extended to the broader data science education community at the scale that they should.
 For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on the accessing data.
 
+In response to the need to develop accessible teaching materials within the data science discipline, this paper demonstrates and shares some teaching strategies that we have employed in our teaching introductory data science courses.
 In this manuscript, we argue that accessibility should be taught early in the data science curriculum so that regardless of whether learners major in data science or not, they can have foundational exposure to accessibility.
+
+Some people may wonder why we need to teach accessibility as part of an introductory data science course as opposed to covering accessibility in a separate and general course module.
+According to the recent faculty survey on teaching accessibility [@shinoharaWhoTeachesAccessibility2018], the most critical barriers to teaching accessibility reported by 1,857 faculty from 318 institutions were "the absence of clear and *discipline-specific accessibility* [emphasis added] learning objectives and the lack of faculty knowledge about accessibility." <!-- Need to find the exact page number for this quote -->
+This means faculty often have trouble making sense of how they can integrate accessibility into their specific teaching domains, which requires applied knowledge of accessibility beyond general concepts.
+
+We share our experiences and strategies in the hope that our discipline-specific (i.e., data science and data visualization) accessibility teaching cases can be used as scaffolding guides for other data science instructors to build upon.
+
+Another critique we may receive includes the timing of when to introduce accessibility concepts in data science courses.
+For example, some instructors may prefer to put teaching accessibility on the back burner so that they can save more time for their students to learn data science fundamentals first and adopt accessibility principles in an upper-level course later.
+While the decision is subject to each institution's and individual instructor's discussion, we believe the sooner, the better.
+This is because: (1) some students may take an introductory level only, which means they will lose a chance to get exposed to this demanding knowledge unless they elect the upper-level course covering accessibility; (2) this kind of separation may unintentionally contribute to misconceptions, such as accessibility is difficult and/or accessibility is optional, not essential.
+When properly incorporated, accessibility can be taught along with other data science fundamentals (e.g., wrangling, merging, tidying, visualizing) while not necessarily putting more cognitive burdens against students' current learning.
+Rather, students can take a chance to become creative design thinkers by making their data science work more meaningful and accessible through "empathy learning" [@putnamTeachingAccessibilityLearning2015].
+Thus, we suggest teaching accessibility as part of an introductory data science course for students to practice coupling their data science knowledge with accessibility principles as early as possible.
+
 As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we share specific examples that can be utilized by other data science instructors.
 We teach R [@R-base] as the main language in our courses; thus, our examples are shaped around using R.
-However, instructors using a different set of language(s) or even teaching language-agnostic courses may still find the content applicable for their own courses.
+However, instructors using a different set of language(s) or even teaching language-agnostic courses may still find the content applicable to their own courses.
 
@@
 Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency [@kimAccessibleVisualizationDesign2021]. 
-Such vision-dependent output may cause critical barriers against blind and visually impaired people and sometimes people with learning disabilities [@kimAccessibleVisualizationDesign2021; @marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
+Such vision-dependent output may cause critical barriers against not only blind and visually impaired people but also  people with cognitive and learning impairments and people having motor impairments [@kimAccessibleVisualizationDesign2021; @marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
 We argue that instructors need to teach multiple data representation methods so that all students can produce data products that are more accessible. 

### 2022-11-07 manuscript.Rmd
add missing sentence to intro

@@
 For example, accessibility has been recently added to engineering design accreditation requirements by Accreditation Board for Engineering and Technology [ABET; abet.org] [@shinoharaWhoTeachesAccessibility2018].
+
+Even though this statistic may be different specifically for data scientists, data science as a highly interdisciplinary field across computing and information sciences also shares the rising demand for accessibility preparedness.
+Furthermore, data science is broad with the intersection in careers such as developers (e.g., data dashboard and package developers) and designers (e.g., data visualization experts).
+Therefore, the impact of teaching and learning accessibility is as far-reaching as its wide scope.
 

### 2022-11-08 manuscript.Rmd
delete line breaks
define AT

@@
 The second learning objective echoes our argument on accessible data representation beyond one-type modality.
-People with disabilities may rely on different types of assistive technology in their day-to-day life.
-Assistive Technology is any form of technology (e.g., software, hardware) that helps people with disabilities perform certain activities.
-Examples of assistive technology include, but are not limited to, walking sticks, wheelchairs, and screen readers.
+People with disabilities may rely on different types of assistive technology in their day-to-day lives.
+Assistive technology (AT) is defined as "products, equipment, and systems that enhance learning, working, and daily living for persons with disabilities" (Assistive Technology Industry Association; [https://bit.ly/3fJN2aJ](https://bit.ly/3fJN2aJ)).
+Examples of assistive technology include, but are not limited to, walking sticks, wheelchairs, screen readers, foot-activated switches, joysticks, sip and puff
+controllers, eye tracking device, and speech recognition input.
 As we noted above, we introduced screen reader as one of these examples when our students were working on a data visualization project; however, we acknowledge that various assistive technologies, such as speech recognition and video captioning, could be integrated into data science courses in diverse contexts (e.g., data collection and data presentation contexts).

### 2022-11-08 manuscript.Rmd
add a missing cite

@@
 While each stage has unique challenges, data visualization may pose fundamental barriers against people with permanent, situational, and temporary disabilities (e.g., visual, cognitive, and motor impairments, to name a few) in accessing information [@marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
-For example, people with cognitive and learning disabilities often have difficulties understanding the abstractions and symbolic conventions used in visualizations, and people lacking fine motor coordination or having tremors may be unable to operate interactive visualizations, not to mention people who are blind and have low vision facing extra challenges in visualization access [@marriottInclusiveDataVisualization2021].
+For example, people with cognitive and learning disabilities often have difficulties understanding the abstractions and symbolic conventions used in visualizations, and people lacking fine motor coordination or having tremors may be unable to operate interactive visualizations, not to mention people who are blind and have low vision facing extra challenges in visualization access [@marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
 @marriottInclusiveDataVisualization2021, in their call to action paper, highlights the serious impact of inaccessible visualizations: "This is an equity issue with severe consequences, because a lack of access to visualizations and underlying data impacts health, educational and work opportunities, and lifestyle." <!-- Need to find a specific page number for this quote.

### 2022-11-08 manuscript.Rmd
better wording

@@
 However, data science is an interdisciplinary field that is taught in various disciplines, including not only computer and information sciences but also statistics, business schools, political science, biological sciences, and more [@schwab2021data].
-Thus, we posit that the resources and conversations around accessibility have not yet extended to the broader data science education community at the scale that they should.
+Therefore, we posit that the resources and conversations around accessibility have not yet extended to the broader data science education community at the scale that they should.
 For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on the accessing data.
@@
 According to the recent faculty survey on teaching accessibility [@shinoharaWhoTeachesAccessibility2018], the most critical barriers to teaching accessibility reported by 1,857 faculty from 318 institutions were "the absence of clear and *discipline-specific accessibility* [emphasis added] learning objectives and the lack of faculty knowledge about accessibility." <!-- Need to find the exact page number for this quote -->
-This means faculty often have trouble making sense of how they can integrate accessibility into their specific teaching domains, which requires applied knowledge of accessibility beyond general concepts.
+This reveals faculty often have trouble making sense of how they can integrate accessibility into their specific teaching domains, which requires applied knowledge of accessibility beyond general concepts.
 

### 2022-11-08 manuscript.Rmd
add other countries' ADA-like act

@@
 As both the authors teach data science in the United States, it was a natural choice for us to make students aware of the Americans with Disabilities Act (ADA).
+However, many countries thesedays may have their own laws similar to ADA, such as Canada’s Accessibility for Ontarians with Disabilities Act, European Accessibility Act, United Kingdom’s Equality Act, Ireland's Equal Status Acts, Australia's Disability Discrimination Act, Israel's Equal Rights of Persons with Disabilities Act, South Korea’s Act on Welfare of Persons with Disabilities, and Japan’s Elimination of Discrimination against Persons with Disabilities Act to name a few.
 The ADA is a civil rights law that prohibits discrimination against individuals with disabilities.

### 2022-11-08 manuscript.Rmd
render Rmd
correct some misspells

--- 2022-11-08 manuscript.Rmd
@@
 For example, people with cognitive and learning disabilities often have difficulties understanding the abstractions and symbolic conventions used in visualizations, and people lacking fine motor coordination or having tremors may be unable to operate interactive visualizations, not to mention people who are blind and have low vision facing extra challenges in visualization access [@marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
-@marriottInclusiveDataVisualization2021, in their call to action paper, highlights the serious impact of inaccessible visualizations: "This is an equity issue with severe consequences, because a lack of access to visualizations and underlying data impacts health, educational and work opportunities, and lifestyle." <!-- Need to find a specific page number for this quote.
+@marriottInclusiveDataVisualization2021, in their call to action paper, highlights the serious impact of inaccessible visualizations: "This is an equity issue with severe consequences because a lack of access to visualizations and underlying data impacts health, educational and work opportunities, and lifestyle." <!-- Need to find a specific page number for this quote.
 -->
@@
 The first objective provides our students with a chance to learn legislative concepts and empathy building concerning diversity issues.
-As both the authors teach data science in the United States, it was a natural choice for us to make students aware of the Americans with Disabilities Act (ADA).
+As both authors teach data science in the United States, it was a natural choice for us to make students aware of the Americans with Disabilities Act (ADA).
+However, many countries these days may have their own laws similar to ADA, such as Canada’s Accessibility for Ontarians with Disabilities Act, European Accessibility Act, United Kingdom’s Equality Act, Ireland's Equal Status Acts, Australia's Disability Discrimination Act, Israel's Equal Rights of Persons with Disabilities Act, South Korea’s Act on Welfare of Persons with Disabilities, and Japan’s Elimination of Discrimination against Persons with Disabilities Act to name a few.
 The ADA is a civil rights law that prohibits discrimination against individuals with disabilities.
-Regardless of their own disability status, as future data scientists, we expect our students to produce data science products that are accessible to all, not only because it is the right thing to do but also because they may have future responsibilities as data scientists to comply to ADA.
+Regardless of their own disability status, as future data scientists, we expect our students to produce data science products that are accessible to all, not only because it is the right thing to do but also because they may have future responsibilities as data scientists to comply with ADA.
 
@@
 Assistive technology (AT) is defined as "products, equipment, and systems that enhance learning, working, and daily living for persons with disabilities" (Assistive Technology Industry Association; [https://bit.ly/3fJN2aJ](https://bit.ly/3fJN2aJ)).
-Examples of assistive technology include, but are not limited to, walking sticks, wheelchairs, screen readers, foot-activated switches, joysticks, sip and puff
-controllers, eye tracking device, and speech recognition input.
+Examples of assistive technology include, but are not limited to, walking sticks, wheelchairs, screen readers, foot-activated switches, joysticks, sip-and-puff
+controllers, eye-tracking devices, and speech recognition input.
 As we noted above, we introduced screen reader as one of these examples when our students were working on a data visualization project; however, we acknowledge that various assistive technologies, such as speech recognition and video captioning, could be integrated into data science courses in diverse contexts (e.g., data collection and data presentation contexts).
@@
 The code creates a grid of four plots where each plot shows a different color vision deficiency, including deuteranomaly, protanomaly, tritanopia, and desaturated.
-This grid allows those without any color vision deficiencies to experience what data visualizations would look like to a sighted person that has color defiency.
+This grid allows those without any color vision deficiencies to experience what data visualizations would look like to a sighted person that has a color deficiency.
 
@@
 The scatterplots are the same as the scatterplot in the previous figure except for colors.
-From left to right and top to bottom the plots read deutanomly, protanomly, tritanomly, and desaturated.
-The deutanomly plot shows two sets of points that are close to brownish green and one set of points that are violet.
-The protanomly plot is similar in colors, one of the brownish green is closer to gray.
-In the tritanomly plot, it is difficult to distinguish the green/blue colors.
+From left to right and top to bottom the plots read deuteranomaly, protanomaly, tritanopia, and desaturated.
+The deuteranomaly plot shows two sets of points that are close to brownish green and one set of points that are violet.
+The protanomaly plot is similar in colors, one of the brownish greens is closer to gray.
+In the tritanopia plot, it is difficult to distinguish the green/blue colors.
 There is also a pink color that stands out.
@@
 For their data visualizations, instead of relying on the default color selection of R packages, students are introduced to the Okabe-Ito color palette, which is known to be accessible to people with color vision deficiencies [@okabe].
-One can find specific colors in the pallette by the R code `palette.colors(palette = "Okabe-Ito")`.
-In addition, students can be introduced to `colorblindr::scale_color_OkabeIto()` and `colorblindr::scale_fill_OkabeIto()` to automatically switch to using this color pallette by adding a layer to a ggplot presentation.
+One can find specific colors in the palette by the R code `palette.colors(palette = "Okabe-Ito")`.
+In addition, students can be introduced to `colorblindr::scale_color_OkabeIto()` and `colorblindr::scale_fill_OkabeIto()` to automatically switch to using this color palette by adding a layer to a ggplot presentation.
 For instance, the plot in Figure \ref{fig:penguins-basic} can be switched to the Okabe-Ito color palette as shown in Figure \ref{fig:penguins-okabe} by using the code below.
 
-```{r penguins-okabe, fig.cap="Scatterplot Using Okabe-Ito Color Pallette", fig.alt="The scatterplots are the same as the scatterplot in the original penguins plot except for colors.
+```{r penguins-okabe, fig.cap="Scatterplot Using Okabe-Ito Color Palette", fig.alt="The scatterplots are the same as the scatterplot in the original penguins' plot except for colors.
 Data points are in orange, sky blue, and bluish green for Adelie, Chinstrap and Gentoo penguins respectively.", warning=FALSE, fig.height=3, echo=TRUE}
@@
 Even though the automated alternate text has some use with its limitations, it does not often directly convey the message a visualization is displaying.
-Thus, it is important to write alternate text manually and teach students how to write one.
+Thus, it is important to write alternate texts manually and teach students how to write one.
 We rely on Cesal's “simple formula for writing alt text for data visualization” [-@cesal], which is as follows: “alt= _Chart type_ of _type of data_ where _reason for including chart_.” 
@@
 
-In this example, a scatter plot is represented by sound. Position on the X axis is communicated using the left-to-right stereo effect; the pitch of the sound indicates the position on the Y axis.
-
+In this example, a scatter plot is represented by sound.
+Position on the X axis is communicated using the left-to-right stereo effect; the pitch of the sound indicates the position on the Y axis.
 Data sonification is very useful when representing linear patterns; however, there is much room for improvement in terms of complex data sonification as of this writing.
@@
 Although we have mainly focused on tools and different representations of data to have students develop accessible content, a more crucial accessibility feature is reproducible workflows. 
-Scholars often focus on reproducibility and replication of scientific findings. 
+Scholars often focus on the reproducibility and replication of scientific findings. 
 We believe reproducibility can often serve as an accessibility feature in a data science context. 

### 2022-11-08 manuscript.Rmd
add more built-in screen reader examples

@@
 Screen reader is an assistive technology that supports blind or visually impaired people in using their computer by reading aloud the contents.
-Recently, screen reader has been provided as one of the built-in accessibility features on operating systems, such as Windows Narrator and Apple's VoiceOver.
+Recently, screen reader has been provided as one of the built-in accessibility features on operating systems, such as Microsoft Windows Narrator, Apple iOS VoiceOver, Android TalkBak and Linux Gnome Orca.
 

### 2022-11-09 manuscript.Rmd
Closes #16

@@
 In the code chunk below, the plot is named as `fig`. 
-The code creates a grid of four plots where each plot shows a different color vision deficiency, including deuteranomaly, protanomaly, tritanopia, and desaturated.
+The code creates a grid of four plots where each plot shows a different color vision deficiency, including deutanomaly, protanomaly, tritanomaly, and desaturated.
 This grid allows those without any color vision deficiencies to experience what data visualizations would look like to a sighted person that has a color deficiency.

### 2022-11-09 manuscript.Rmd
fix typos in intro

@@
 Therefore, we posit that the resources and conversations around accessibility have not yet extended to the broader data science education community at the scale that they should.
-For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on the accessing data.
+For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on accessing the data.
 

### 2022-11-09 manuscript.Rmd
fix typos in intro

@@
 
-Despite the increasing demand for accessibility knowledge in both industries and academia and the critical impact of accessible visualizations, accessibility has not yet found its place in the data science course and teaching materials.
+Despite the increasing demand for accessibility knowledge in both industry and academia and the critical impact of accessible visualizations, accessibility has not yet found its place in the data science course and teaching materials.
 One main reason for this could be that accessibility is often incorporated into human-computer interaction (HCI) and design courses [@shinoharaHowDesigningPeople2016; @shinoharaTeachingAccessibilityTechnology2017a], introductory web development [@koAccessComputingPromotesTeaching2016; @rosmaitaAccessibilityFirstNew2006], software engineering courses [@martin-escalonaIntroductionTopicAccessibility2013; @ludiIntroducingAccessibilityRequirements2007; @ludiAccessEveryoneIntroducing2002] or the digital arts [@barata2019inclusion], if at all covered.
@@
 
-Another critique we may receive includes the timing of when to introduce accessibility concepts in data science courses.
+Another important aspect in curriculum design includes the timing of when to introduce accessibility concepts in data science courses.
 For example, some instructors may prefer to put teaching accessibility on the back burner so that they can save more time for their students to learn data science fundamentals first and adopt accessibility principles in an upper-level course later.
@@
 This is because: (1) some students may take an introductory level only, which means they will lose a chance to get exposed to this demanding knowledge unless they elect the upper-level course covering accessibility; (2) this kind of separation may unintentionally contribute to misconceptions, such as accessibility is difficult and/or accessibility is optional, not essential.
-When properly incorporated, accessibility can be taught along with other data science fundamentals (e.g., wrangling, merging, tidying, visualizing) while not necessarily putting more cognitive burdens against students' current learning.
+When properly incorporated, accessibility can be taught along with other data science fundamentals (e.g., wrangling, merging, tidying, visualizing) while not necessarily putting more cognitive burden against students' current learning.
 Rather, students can take a chance to become creative design thinkers by making their data science work more meaningful and accessible through "empathy learning" [@putnamTeachingAccessibilityLearning2015].

### 2022-11-09 manuscript.Rmd
fix a misspelling

@@
 In the code chunk below, the plot is named as `fig`. 
-The code creates a grid of four plots where each plot shows a different color vision deficiency, including deutanomaly, protanomaly, tritanomaly, and desaturated.
+The code creates a grid of four plots where each plot shows a different color vision deficiency, including deuteranomaly, protanomaly, tritanomaly, and desaturated.
 This grid allows those without any color vision deficiencies to experience what data visualizations would look like to a sighted person that has a color deficiency.
@@
 The scatterplots are the same as the scatterplot in the previous figure except for colors.
-From left to right and top to bottom the plots read deuteranomaly, protanomaly, tritanopia, and desaturated.
+From left to right and top to bottom the plots read deuteranomaly, protanomaly, tritanomaly, and desaturated.
 The deuteranomaly plot shows two sets of points that are close to brownish green and one set of points that are violet.
 The protanomaly plot is similar in colors, one of the brownish greens is closer to gray.
-In the tritanopia plot, it is difficult to distinguish the green/blue colors.
+In the tritanomaly plot, it is difficult to distinguish the green/blue colors.
 There is also a pink color that stands out.

### 2022-11-09 manuscript.Rmd
add page number

@@
 Some people may wonder why we need to teach accessibility as part of an introductory data science course as opposed to covering accessibility in a separate and general course module.
-According to the recent faculty survey on teaching accessibility [@shinoharaWhoTeachesAccessibility2018], the most critical barriers to teaching accessibility reported by 1,857 faculty from 318 institutions were "the absence of clear and *discipline-specific accessibility* [emphasis added] learning objectives and the lack of faculty knowledge about accessibility." <!-- Need to find the exact page number for this quote -->
+According to the recent faculty survey on teaching accessibility [@shinoharaWhoTeachesAccessibility2018], the most critical barriers to teaching accessibility reported by 1,857 faculty from 318 institutions were "the absence of clear and *discipline-specific accessibility* [emphasis added] learning objectives and the lack of faculty knowledge about accessibility" (p.
+197).
 This reveals faculty often have trouble making sense of how they can integrate accessibility into their specific teaching domains, which requires applied knowledge of accessibility beyond general concepts.

### 2022-11-09 manuscript.Rmd
add missing page numbers

@@
 For example, people with cognitive and learning disabilities often have difficulties understanding the abstractions and symbolic conventions used in visualizations, and people lacking fine motor coordination or having tremors may be unable to operate interactive visualizations, not to mention people who are blind and have low vision facing extra challenges in visualization access [@marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
-@marriottInclusiveDataVisualization2021, in their call to action paper, highlights the serious impact of inaccessible visualizations: "This is an equity issue with severe consequences because a lack of access to visualizations and underlying data impacts health, educational and work opportunities, and lifestyle." <!-- Need to find a specific page number for this quote.
--->
+@marriottInclusiveDataVisualization2021, in their call to action paper, highlights the serious impact of inaccessible visualizations: "This is an equity issue with severe consequences because a lack of access to visualizations and underlying data impacts health, educational and work opportunities, and lifestyle" (p.
+48).
 

### 2022-11-13 manuscript.Rmd
address reviewer comments

@@
 
-Another important aspect in curriculum design includes the timing of when to introduce accessibility concepts in data science courses.
+Another important aspect of curriculum design includes the timing of when to introduce accessibility concepts in data science courses.
 For example, some instructors may prefer to put teaching accessibility on the back burner so that they can save more time for their students to learn data science fundamentals first and adopt accessibility principles in an upper-level course later.
@@
 This is because: (1) some students may take an introductory level only, which means they will lose a chance to get exposed to this demanding knowledge unless they elect the upper-level course covering accessibility; (2) this kind of separation may unintentionally contribute to misconceptions, such as accessibility is difficult and/or accessibility is optional, not essential.
-When properly incorporated, accessibility can be taught along with other data science fundamentals (e.g., wrangling, merging, tidying, visualizing) while not necessarily putting more cognitive burden against students' current learning.
+When properly incorporated, accessibility can be taught along with other data science fundamentals (e.g., wrangling, merging, tidying, visualizing) while not necessarily putting more cognitive burden on students' current learning.
 Rather, students can take a chance to become creative design thinkers by making their data science work more meaningful and accessible through "empathy learning" [@putnamTeachingAccessibilityLearning2015].
@@
 We believe that a natural connection between data science and accessibility can be achieved in the data science classroom while presenting data visualization and then introducing the other aforementioned forms of data representation.
-While delivering the content, instructors should try to provide as many opportunities as possible for students to experience using these tools themselves. 
-For instance, each student should use a screen reader at least once regardless of their visual ability status and hear the alternate text that they have written.
+While delivering the content, instructors should try to provide as many opportunities as possible for students to experience using these tools and accessibility principles themselves. 
+For instance, each student could use screen reader at least once regardless of their visual acuity and hear the alternate text that they have written.
 
 Ample opportunities in learning accessibility should also include assessments. 
-For instance, instructors incorporating accessibility should modify assessment instructions and rubrics to include accessibility. 
+For instance, instructors should modify assessment instructions and rubrics to include accessibility. 
 This will prevent students from just hearing about accessibility in the lecture and then forgetting about it. 
 Ideally, the assessments should not treat accessibility as a learning objective at a single point in the academic term. 
-For instance, if accessibility is covered in Week 3 of the term, students can (and should) still be expected to write alternate texts for visualizations in Week 10.
+For instance, if accessibility is covered in Week 3 of the term, students can (and should) still be expected to write alternate texts for visualizations until the end of the semester.
 
@@
 We believe reproducibility can often serve as an accessibility feature in a data science context. 
-A reproducible open-source workflow allows users who are blind or have a visual impairment to examine data in their preferred form of data representation.
+A reproducible open-source workflow, for example, allows users with and without dis/abilities to examine data in their preferred form of data representation.
 Thus, in addition to teaching different representations of data, instructors may also consider teaching reproducibility.
@@
 We also acknowledge that what we have presented here is only a small portion of the intersection of visual accessibility and data science.
+
+Thus far, we have just demonstrated some methods to teach accessibility in introductory data science courses with a special focus on accessible visualization.
+However, as @shinoharaWhoTeachesAccessibility2018 highlight, teaching accessibility at scale requires large-scale adoption and implementation of these methods by faculty, who are organically contributing to, and constructing, instructors’ knowledge [@barnettConceptualisingCurriculumChange2001; @kirkTeacherVoiceOwnership2001].
+We hope that this manuscript has just opened the discussion on teaching accessibility in the data science field, and we encourage other data science faculty to help us move towards large-scale curricular change beyond our examples.
+
+# Acknowledgements
+
 The second co-author was supported through a Teach Access grant to develop curricular materials on teaching accessibility as part of the data science curriculum. 
 Educators interested in teaching accessibility can follow the organization's work at https://teachaccess.org/.
+
+# Supplementary Materials
+
 The reproducible R source code of this manuscript can be found on a GitHub repository at \url{\repo{}}.

### 2022-11-13 revision_letter.Rmd
address reviewer comments

--- 2022-11-03 revision_letter.Rmd
@@
 class       : "draft"
-output      : papaja::revision_letter_pdf
-bibliography: ["bib/packages.bib", "bib/references.bib"]
+output:
+  papaja::revision_letter_pdf:
+    keep_tex: true
+bibliography: ["bib/packages.bib", "bib/references.bib", "bib/teachinga11y.bib"]
+csl: "bib/chicago-author-date.csl"
 ---
@@
 
-# AE
+# AE {#response1}
 
@@
 
-This is our response
+We appreciate your thoughtful comment. In response to this feedback, we have elaborated our argument more clearly in our manuscript.
 
-\Assignment{First Author}
-\WorkInProgress
-\Easy
+The following includes part of the `Introduction` section quoted from the revised manuscript to illustrate the change:
+
+> Some people may wonder why we need to teach accessibility as part of an introductory data science course as opposed to covering accessibility in a separate and general course module.
+According to the recent faculty survey on teaching accessibility [@shinoharaWhoTeachesAccessibility2018], the most critical barriers to teaching accessibility reported by 1,857 faculty from 318 institutions were "the absence of clear and *discipline-specific accessibility* [emphasis added] learning objectives and the lack of faculty knowledge about accessibility" (p.
+197).
+This reveals faculty often have trouble making sense of how they can integrate accessibility into their specific teaching domains, which requires applied knowledge of accessibility beyond general concepts.
+> . . .
+> Another important aspect of curriculum design includes the timing of when to introduce accessibility concepts in data science courses.
+For example, some instructors may prefer to put teaching accessibility on the back burner so that they can save more time for their students to learn data science fundamentals first and adopt accessibility principles in an upper-level course later.
+While the decision is subject to each institution's and individual instructor's discussion, we believe the sooner, the better.
+This is because: (1) some students may take an introductory level only, which means they will lose a chance to get exposed to this demanding knowledge unless they elect the upper-level course covering accessibility; (2) this kind of separation may unintentionally contribute to misconceptions, such as accessibility is difficult and/or accessibility is optional, not essential.
+When properly incorporated, accessibility can be taught along with other data science fundamentals (e.g., wrangling, merging, tidying, visualizing) while not necessarily putting more cognitive burden on students' current learning.
+Rather, students can take a chance to become creative design thinkers by making their data science work more meaningful and accessible through "empathy learning" [@putnamTeachingAccessibilityLearning2015].
+Thus, we suggest teaching accessibility as part of an introductory data science course for students to practice coupling their data science knowledge with accessibility principles as early as possible.
 
@@
 
-\Assignment{Second Author}
-\AlmostDone
-\Medium
-
 \RC{
@@
 
-This is our response: need to create a section dededicated to a11y, and say that we are presenting this paper as a demonstration how we integrated a11y into our courses.
+Thank you for the constructive feedback. We strengthen our argument by adding the following paragraphs in the `Introduction` section.
 
-\Assignment{First Author}
-\Done
-\TimeConsuming
-\Hard
-\NotEstimated
+> Data science consists of an iterative process of importing, tidying, transforming, visualizing, and modeling data [@wickham2016r].
+While each stage has unique challenges, data visualization may pose fundamental barriers against people with permanent, situational, and temporary disabilities (e.g., visual, cognitive, and motor impairments, to name a few) in accessing information [@marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
+For example, people with cognitive and learning disabilities often have difficulties understanding the abstractions and symbolic conventions used in visualizations, and people lacking fine motor coordination or having tremors may be unable to operate interactive visualizations, not to mention people who are blind and have low vision facing extra challenges in visualization access [@marriottInclusiveDataVisualization2021; @leeReachingBroaderAudiences2020].
+@marriottInclusiveDataVisualization2021, in their call to action paper, highlights the serious impact of inaccessible visualizations: "This is an equity issue with severe consequences because a lack of access to visualizations and underlying data impacts health, educational and work opportunities, and lifestyle" (p.
+48).
+> Despite the increasing demand for accessibility knowledge in both industry and academia and the critical impact of accessible visualizations, accessibility has not yet found its place in the data science course and teaching materials.
+One main reason for this could be that accessibility is often incorporated into human-computer interaction (HCI) and design courses [@shinoharaHowDesigningPeople2016; @shinoharaTeachingAccessibilityTechnology2017a], introductory web development [@koAccessComputingPromotesTeaching2016; @rosmaitaAccessibilityFirstNew2006], software engineering courses [@martin-escalonaIntroductionTopicAccessibility2013; @ludiIntroducingAccessibilityRequirements2007; @ludiAccessEveryoneIntroducing2002] or the digital arts [@barata2019inclusion], if at all covered.
+However, data science is an interdisciplinary field that is taught in various disciplines, including not only computer and information sciences but also statistics, business schools, political science, biological sciences, and more [@schwab2021data].
+Therefore, we posit that the resources and conversations around accessibility have not yet extended to the broader data science education community at the scale that they should.
+For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on accessing the data.
+> In response to the need to develop accessible teaching materials within the data science discipline, this paper demonstrates and shares some teaching strategies that we have employed in our teaching introductory data science courses.
 
@@
 
-This is our response
-
-\Assignment{First Author}
-\WorkInProgress
-\Easy
+Thanks for the insightful comment. We have addressed this question as described in our responses to AE's comments above.
 
@@
 
-This is our response: good point so it is critical to share the same principle across different and with upper level courses .
+Thank you very much for your great input.
+Although we demonstrated our examples in the context of introductory data science courses, we believe accessibility needs to be taught in other data science courses, including higher levels.
+This requires instructors' broader efforts.
+We have clarified this in our `Closing Remarks` section like below:
 
-\Assignment{First Author}
-\WorkInProgress
-\Easy
+> Thus far, we have just demonstrated some methods to teach accessibility in introductory data science courses with a special focus on accessible visualization.
+> However, as @shinoharaWhoTeachesAccessibility2018 highlight, teaching accessibility at scale requires large-scale adoption and implementation of these methods by faculty, who are organically contributing to, and constructing, instructors’ knowledge [@barnettConceptualisingCurriculumChange2001; @kirkTeacherVoiceOwnership2001].
+We hope that this manuscript has just opened the discussion on teaching accessibility in the data science field, and we encourage other data science faculty to help us move towards large-scale curricular change beyond our examples.
 
@@
 
-This is our response
-
-\Assignment{First Author}
-\WorkInProgress
-\Easy
+We have addressed this comment above via [our first response to AE's comment](#response1).
 

### 2022-11-14 manuscript.Rmd
extend thanks to reviewers and AE, separate paragraphs for easier readibility

@@
 
-Some people may wonder why we need to teach accessibility as part of an introductory data science course as opposed to covering accessibility in a separate and general course module.
+One may wonder why we need to teach accessibility as part of an introductory data science course as opposed to covering accessibility in a separate and general course module.
 According to the recent faculty survey on teaching accessibility [@shinoharaWhoTeachesAccessibility2018], the most critical barriers to teaching accessibility reported by 1,857 faculty from 318 institutions were "the absence of clear and *discipline-specific accessibility* [emphasis added] learning objectives and the lack of faculty knowledge about accessibility" (p.

### 2022-11-14 revision_letter.Rmd
extend thanks to reviewers and AE, separate paragraphs for easier readibility

@@
 
-Thank you very much for taking the time to consider our manuscript for publication at _`r rmarkdown::metadata$journal`_.
-In the following we address your and each reviewers' concerns point-by-point.
+Thank you, the associate editor, and the reviewers for taking the time to consider our manuscript for publication at _`r rmarkdown::metadata$journal`_.
+The constructive comments have helped us revise and improve the manuscript.
+In the following we address each of your concerns point-by-point.
 
@@
 
-The following includes part of the `Introduction` section quoted from the revised manuscript to illustrate the change:
+We quote below two separate parts that we have added to the `Introduction` section.
 
-> Some people may wonder why we need to teach accessibility as part of an introductory data science course as opposed to covering accessibility in a separate and general course module.
+> One may wonder why we need to teach accessibility as part of an introductory data science course as opposed to covering accessibility in a separate and general course module.
 According to the recent faculty survey on teaching accessibility [@shinoharaWhoTeachesAccessibility2018], the most critical barriers to teaching accessibility reported by 1,857 faculty from 318 institutions were "the absence of clear and *discipline-specific accessibility* [emphasis added] learning objectives and the lack of faculty knowledge about accessibility" (p.
@@
 This reveals faculty often have trouble making sense of how they can integrate accessibility into their specific teaching domains, which requires applied knowledge of accessibility beyond general concepts.
-> . . .
+
 > Another important aspect of curriculum design includes the timing of when to introduce accessibility concepts in data science courses.
@@
 48).
+
 > Despite the increasing demand for accessibility knowledge in both industry and academia and the critical impact of accessible visualizations, accessibility has not yet found its place in the data science course and teaching materials.
@@
 For instance, many introductory data science courses [@yan2019first; @baumer2015data], textbooks [@adhikari2019computational; @wickham2016r], or curricula [@schwab2021data] either fail to mention accessibility at all or only focus on accessing the data.
+
 > In response to the need to develop accessible teaching materials within the data science discipline, this paper demonstrates and shares some teaching strategies that we have employed in our teaching introductory data science courses.

### 2022-11-14 revision_letter.Rmd
Closes #17

@@
 
-This is our response: For example, instructors can include accessibility as one of their evaluation criteria for course assignments or make it as a chance to obtain bonus credits for the students' visualization work.
+We have clarified this point with the following sentence, in addition to the original sentence in the `Closing Remarks` section.
 
-> This is a section quoted from the revised manuscript to illustrate the change.
+> We believe that an accessible future in data science requires educators to teach accessibility as part of the curriculum deliberately. 
+This includes both teaching the topic as well as assessing it so that students are held accountable for their accessibility practices.
+
+Please note that we had also the following section in the original manuscript which we have also kept in the revised version.
+
+> Ample opportunities in learning accessibility should also include assessments. 
+For instance, instructors should modify assessment instructions and rubrics to include accessibility. 
+This will prevent students from just hearing about accessibility in the lecture and then forgetting about it. 
+Ideally, the assessments should not treat accessibility as a learning objective at a single point in the academic term. 
+For instance, if accessibility is covered in Week 3 of the term, students can (and should) still be expected to write alternate texts for visualizations until the end of the semester.
 

### 2022-11-14 manuscript.Rmd
Closes #17
render Rmd files

@@
 We believe that an accessible future in data science requires educators to teach accessibility as part of the curriculum deliberately. 
+This includes both teaching the topic as well as assessing it so that students are held accountable for their accessibility practices.
 We also acknowledge that what we have presented here is only a small portion of the intersection of visual accessibility and data science.
@@
 ```
-
-# Lit Review
-
-## @shinoharaWhoTeachesAccessibility2018
-
-### RP
-
-Industry demand for software developers with knowledge of accessibility has increased substantially in recent years.
-However, there is little knowledge about the prevalence of higher education teaching about accessibility or faculty's perceived barriers to teaching accessibility.
-
-### RQ
-
-* RQ1. Who is teaching accessibility?
-* RQ2. What barriers do faculty see to teaching accessibility?
-* RQ3. What factors predict who is teaching accessibility?
-
-### DC
-
-surveyed 14,176 computing and information science faculty in the United States.
-We received a representative sample of at least one response from 318 of the 352 institutions we surveyed, totaling 1,857 responses.
-
-### Findings
-
-175 institutions (50%) had at least one instructor teaching accessibility and that no fewer than 2.5% of faculty overall teach accessibility.
-Faculty that teach accessibility are twice as likely to be female, to have expertise in HCI and software engineering, and to know people with disabilities.
-
-Overall, 20% of faculty (375) responded “Yes” to the prompt “Do you teach courses that incorporate topics about accessibility?”
-
-The most critical barriers to teaching accessibility that faculty reported were the absence of clear and discipline-specific accessibility learning objectives and the lack of faculty knowledge about accessibility.
-
-Faculty desired resources that were specific to the areas of computing in which they teach rather than general accessibility resources and guidelines.
-
-### Quotes
-
-<!-- Increasing interest and importance of accessibility -->
-> Consortia like TeachAccess (teachaccess.org), led by Yahoo!, Facebook, Microsoft, Google, and Adobe, represent this increased demand for accessibility knowledge, and are leading efforts to increase the number of computer and information science faculty teaching accessibility.
-
-> In parallel, professional societies such as ABET [abet.org] have recently added accessibility into engineering design accreditation requirements.
-
-<!-- Examples of how accessibility is integrated into computing and information sciences courses -->
-> some have incorporated accessibility in software engineering courses [@martin-escalonaIntroductionTopicAccessibility2013; @ludiIntroducingAccessibilityRequirements2007; @ludiAccessEveryoneIntroducing2002], as part of introducing web development [@koAccessComputingPromotesTeaching2016; @rosmaitaAccessibilityFirstNew2006], or in HCI and design courses as a central component of design process [@shinoharaHowDesigningPeople2016; @shinoharaTeachingAccessibilityTechnology2017a].
-
-<!-- Two efforts are required to make broader impacts for teaching accessibility -->
-> Incorporating instruction on accessibility at scale requires two major elements: (1) methods for how to teach accessibility in computer science (CS) and related fields, and (2) large scale adoption and implementation of these methods by faculty.
-<!-- In conclusion, we have just demonstrated some methods to make data science visualization more accessible, focusing on the second element.
-However, we need more broad effort to make it at a large scale as @shinoharaWhoTeachesAccessibility2018 said.
--->
-
-> These works suggest that large scale curricular change is not a matter of top-down decree, but one of fostering local change agents who are intimately aware of the culture of specific institutions and who can devise and implement strategies to work within them.
-
-<!-- Active culture movement is required towards teaching accessibility -->
-> theories of curricular change argue that change only comes through social and instructional discourse among faculty organically contributing to, and constructing, instructors’ knowledge [@barnettConceptualisingCurriculumChange2001; @kirkTeacherVoiceOwnership2001].
-
-> Kezar and Eckel, in an examination of several institutions, determined that institutional culture could enhance or stymie curriculum changes [@kezarEffectInstitutionalCulture2002].
-
-## @kawasTeachingAccessibilityDesign2019
-
-### RP
-
-Most CS students learn little about accessibility in higher education; this is partly because most CS faculty know little about accessibility.
-
-higher education CS faculty lack a model of professional development for learning to teach new topics.
-
-### DC
-
-conducted 18 semi-structured interviews with U.S. CS faculty, asking them to explore a prototype of a web-based professional development tool that linked accessibility topics to CS topics.
-
-we investigated the feasibility of a "micro" professional development model for teaching accessibility in CS courses that could be used at scale.
-
-### Findings
-
-many organizational factors limited faculty's autonomy to integrate accessibility in many of their courses.
-
-individual values and knowledge constrained faculty's ability and willingness to both learn and integrate accessibility topics into their courses.
-
-However, many faculty expressed desire to teach accessibility in their courses if they had access to even basic accessibility content and materials to use in their courses.
-
-## @marriottInclusiveDataVisualization2021
-
-### Quotes
-
-<!-- how many people with disabilities in the world? -->
-> A 2011 report from the WHO (https://bit.ly/3raT0kg) estimates that more than 1 billion people—15 percent of the world’s population— live with some form of disability, of whom about one-fifth experience significant difficulties in functioning.
-While it may be obvious that people who are blind or have low vision (BLV) would face challenges in accessing data visualizations, this also holds for other kinds of disability.
-People who have cognitive and learning disabilities (CALD) frequently have trouble making sense of the abstractions and symbolic conventions used in visualizations, and people who lack fine motor coordination or have tremors may be unable to operate interactive visualizations.
-
-<!-- Three types of major disabilities  affected by inaccessible viz -->
-> There are three main kinds of disability that may affect access to visualizations: visual, cognitive, and motor impairments. It is not uncommon for people to have more than one kind of impairment.
-
-<!-- Why is accessible Visualization important? -->
-> this is an equity issue with severe consequences, because a lack of access to visualizations and underlying data impacts health, educational and work opportunities, and lifestyle.
-For instance, the BLV community has identified a lack of access to educational materials [@butlerUnderstandingGraphicalChallenges2017] and online data visualizations about Covid-19 as a significant concern (https://bit.ly/2NJzy0C).
-
-<!-- People are not paying enough attention to accessible viz -->
-> The visualization research community, however, has not paid enough attention to the needs of people with disabilities [@leeReachingBroaderAudiences2020].
-
-<!-- Accessible viz for all beyond disabilities -->
-> Inclusive approaches to visualizations are likely to help more than those people with permanent disabilities.
-There are also hundreds of millions of people with temporary or situational disabilities.
-New, inclusive approaches could readily become “informational curb cuts” that benefit everyone.
-
-<!-- Tactile representation for blind people -->
-> The Braille Authority of North America (BANA; http://www.
-brailleauthority.org) recommends the use of tactile graphics to show spatial data.
-These are usually raised line drawings, but more recently, 3D-printed models have been used.
-Research suggests that despite the differences between tactile and visual perception, spatial perception and reasoning is fundamental to the use of both visual and tactile graphics, and that tactile maps, scatter plots, and network diagrams offer similar benefits to the corresponding visual graphic (e.g., @yangTactilePresentationNetwork2020a).
-
-<!-- Tactile production is expensive -->
-> Unfortunately, the cost and time for the production of tactile graphics limits their use for data exploration.
-More recently, researchers have explored the multimodal presentation of graphics on touchscreen devices, providing audio and haptic feedback when the reader’s fingers touch the graphic.
-There is considerable opportunity to explore data-exploration tools leveraging refreshable tactile displays as well as touchscreen-based presentations, as these may provide similar benefits to tactile graphics.
-
-<!-- critical disability studies -->
-@mankoffDisabilityStudiesSource2010
-
-<!-- The following is good to review when having more time -->
-
-## ## @joynerVisualizationAccessibilityWild2022
-
-### rp
-
-### rq
-
-### dc
-
-### ### findings
-
-## @koAccessComputingPromotesTeaching2016
-
-### RP
-
-### RQ
-
-### DC
-
-### Findings

### 2022-11-16 manuscript.Rmd
change acm to american computing machinery

@@
 However, a consensus on what should be included in such curricula has not yet been reached.
-One of the few guidelines on the topic is written by the ACM Data Science Task Force and titled *Computing Competencies for Undergraduate Data Science Curricula*.
+One of the few guidelines on the topic is written by the American Computing Machinery Data Science Task Force and titled *Computing Competencies for Undergraduate Data Science Curricula*.
 In this report, accessibility is a "foundational consideration" at the knowledge level when it comes to displaying data [@danyluk2021computing].

### 2022-11-16 manuscript.Rmd
a sentence was by itself all alone

@@
 
-However, we also want to make our teaching as inclusive as possible to other types of dis/abilities so that we are not just confined to issues of visual disabilities.
-
-Therefore, our curricular goal is to broadly ensure that the current and the next generation of data scientists provide public-facing outputs, including websites and analysis reports, in accessible forms and representations for people with diverse abilities.
+However, we also want to make our teaching as inclusive as possible to other types of dis/abilities so that we are not just confined to issues of visual disabilities.Therefore, our curricular goal is to broadly ensure that the current and the next generation of data scientists provide public-facing outputs, including websites and analysis reports, in accessible forms and representations for people with diverse abilities.
 To achieve this goal, we set three specific and minimal learning objectives like below.

### 2022-11-16 manuscript.Rmd
cosmetic change and add missing space

--- 2022-11-16 manuscript.Rmd
@@
 
-However, we also want to make our teaching as inclusive as possible to other types of dis/abilities so that we are not just confined to issues of visual disabilities.
-
+However, we also want to make our teaching as inclusive as possible to other types of dis/abilities so that we are not just confined to issues of visual impairments.
 Therefore, our curricular goal is to broadly ensure that the current and the next generation of data scientists provide public-facing outputs, including websites and analysis reports, in accessible forms and representations for people with diverse abilities.

### 2023-02-24 revision_letter2.Rmd
add second round of response to editor

---
title       : "Teaching Visual Accessibility in Introductory Data Science Classes with Multi-Modal Data Representations"
authors     : "JooYoung Seo & Mine Dogucu"
journal     : "Journal of Data Science"
manuscript  : "JDS2207-022"

class       : "draft"
output:
  papaja::revision_letter_pdf:
    keep_tex: true
bibliography: ["bib/packages.bib", "bib/references.bib", "bib/teachinga11y.bib"]
csl: "bib/chicago-author-date.csl"
---

Dear Dr. Claire McKay Bowen,

Thank you, the associate editor, and the reviewer for taking the time to consider our manuscript for publication at _`r rmarkdown::metadata$journal`_.

We found the feedback to be supportive and helpful, however, we were not able to make the requested minor revisions. We outline our reasoning below.

# Comment

\RC{
    I agree with the AE and reviewer that the manuscript is strong. Before accepting the paper, if possible, we would like to see example assignments, course outlines, and/or feedback from students and instructors when incorporating your methods into your courses. This would provide other instructors a proof of concept when integrating your methods to their courses.
    }

We appreciate the support for the strength of the manuscript. We respond to the requested changes in three separate parts.

**Course outlines**

There are two approaches to teaching accessibility (in data science)

1. As a full course solely focused on accessibility (usually taught in information and computer sciences)
2. As part of other courses (e.g. in introductory data science in our case)

We believe a course outline would make sense to share if we were teaching a full course on accessibility. 
However, we are teaching it similar to the latter option and this is what we cover in the manuscript.
Thus a course outline we provide would be an outline for an introductory data science which would in a way be irrelevant in the scope of this manuscript.
Afterall, we are not discussing whether to teach topics such as web scraping, working with dates, linear regression or not in an introductory course, which all are part of our course outlines and beyond the scope of the manuscript.

Since we are teaching accessibility as part of a course on data science, we had tried to answer the question "when to teach accessibility" in the manuscript already so that intructors would know whether to include it in their Week 2 content or Week 10 content.
We had stated "We believe that a natural connection between data science and accessibility can be achieved in the data science classroom while presenting data visualization and then introducing the other aforementioned forms of data representation."
Thus rather than providing a course outline, we are guiding instructors to include accessibility along with data visualization wherever that may fall in their own course outlines.

**Example Assignments**

Similar to the reasons for course outlines, we do not have any assignments specifically assessing accessibility.
We had, however, already provided points about assessment in the manuscript as follows "For instance,
instructors incorporating accessibility should modify assessment instructions and rubrics to include
accessibility. This will prevent students from just hearing about accessibility in the lecture
and then forgetting about it. Ideally, the assessments should not treat accessibility as a learning
objective at a single point in the academic term. For instance, if accessibility is covered in Week 3
of the term, students can (and should) still be expected to write alternate texts for visualizations
in Week 10."

**Feedback from students and instructors**

Any such data, whether in anecdotal form or course evaluation format, would require an approval from Intitutional Review Board (IRB). 
The second author was supported by a grant to develop part of the teaching materials mentioned in the manuscript. 
They collected data for grant evaluation purposes. 
However, since the data are collected for evaluation purposes and not research purposes, they cannot be used or shared for research purposes without IRB approval. 
Given the time line of the special issue, we will not be seeking IRB approval but hope to be able to publish with empirical data evidence in the future.

### 2023-02-24 revision_letter2.Rmd
made cosmetic changes

@@
 
-Thank you, the associate editor, and the reviewer for taking the time to consider our manuscript for publication at _`r rmarkdown::metadata$journal`_.
-
-We found the feedback to be supportive and helpful, however, we were not able to make the requested minor revisions. We outline our reasoning below.
+We are thankful for your time and consideration of our manuscript to be published in `r rmarkdown::metadata$journal`. Your comments were insightful and we
+appreciate the feedback you have provided.
+Unfortunately, despite our best efforts, we may not be able to make the requested minor revisions due to constraints on our part. Our reasoning is explained
+further below.
 
@@
 We believe a course outline would make sense to share if we were teaching a full course on accessibility. 
-However, we are teaching it similar to the latter option and this is what we cover in the manuscript.
-Thus a course outline we provide would be an outline for an introductory data science which would in a way be irrelevant in the scope of this manuscript.
-Afterall, we are not discussing whether to teach topics such as web scraping, working with dates, linear regression or not in an introductory course, which all are part of our course outlines and beyond the scope of the manuscript.
+However, we are teaching it similarly to the latter option and this is what we cover in the manuscript.
+Thus a course outline we provide would be an outline for introductory data science which would in a way be irrelevant in the scope of this manuscript.
+After all, we are not discussing whether to teach topics such as web scraping, working with dates, and linear regression or not in an introductory course, which all are part of our course outlines and beyond the scope of the manuscript.
 
-Since we are teaching accessibility as part of a course on data science, we had tried to answer the question "when to teach accessibility" in the manuscript already so that intructors would know whether to include it in their Week 2 content or Week 10 content.
-We had stated "We believe that a natural connection between data science and accessibility can be achieved in the data science classroom while presenting data visualization and then introducing the other aforementioned forms of data representation."
+Since we are teaching accessibility as part of a course on data science, we had tried to answer the question "when to teach accessibility" in the manuscript already so that instructors would know whether to include it in their Week 2 content or Week 10 content.
+We had stated, "We believe that a natural connection between data science and accessibility can be achieved in the data science classroom while presenting data visualization and then introducing the other aforementioned forms of data representation."
 Thus rather than providing a course outline, we are guiding instructors to include accessibility along with data visualization wherever that may fall in their own course outlines.
@@
 
-Any such data, whether in anecdotal form or course evaluation format, would require an approval from Intitutional Review Board (IRB). 
+Any such data, whether in anecdotal form or course evaluation format, would require approval from Intitutional Review Board (IRB). 
 The second author was supported by a grant to develop part of the teaching materials mentioned in the manuscript. 
@@
 However, since the data are collected for evaluation purposes and not research purposes, they cannot be used or shared for research purposes without IRB approval. 
-Given the time line of the special issue, we will not be seeking IRB approval but hope to be able to publish with empirical data evidence in the future.
+Given the timeline of the special issue, we will not be seeking IRB approval but hope to be able to publish empirical data evidence in the future.
